### 2016-08-09 Paper/proceedings.tex
Added CHI template.

\documentclass{sigchi}

% Use this command to override the default ACM copyright statement
% (e.g. for preprints).  Consult the conference website for the
% camera-ready copyright statement.

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
% \toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is      granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

% Arabic page numbers for submission.  Remove this line to eliminate
% page numbers for the camera ready copy
% \pagenumbering{arabic}

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{mathptmx}
\usepackage[pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}
% Some optional stuff you might like/need.
\usepackage{microtype} % Improved Tracking and Kerning
% \usepackage[all]{hypcap}  % Fixes bug in hyperref caption linking
\usepackage{ccicons}  % Cite your images correctly!
% \usepackage[utf8]{inputenc} % for a UTF8 editor only

% If you want to use todo notes, marginpars etc. during creation of your draft document, you
% have to enable the "chi_draft" option for the document class. To do this, change the very first
% line to: "\documentclass[chi_draft]{sigchi}". You can then place todo notes by using the "\todo{...}"
% command. Make sure to disable the draft option again before submitting your final document.
\usepackage{todonotes}

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Visual Estimation of Trend in Bivariate Visualizations}
\def\plainauthor{Removed for Review}
\def\emptyauthor{}
\def\plainkeywords{Information Visualization, Graphical Perception, Regression}
\def\plaingeneralterms{Visualization}

% llt: Define a global style for URLs, rather that the default one
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{
    \def\UrlFont{\sf}
  }{
    \def\UrlFont{\small\bf\ttfamily}
  }}
\makeatother
\urlstyle{leo}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={\plaintitle},
% Use \plainauthor for final version.
%  pdfauthor={\plainauthor},
  pdfauthor={\emptyauthor},
  pdfkeywords={\plainkeywords},
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
}

% create a shortcut to typeset table headings
% \newcommand\tabhead[1]{\small\textbf{#1}}

% End of preamble. Here it comes the document.
\begin{document}

\title{\plaintitle}

\numberofauthors{2}
\author{%
%  \alignauthor{Michael Correll\\
%    \affaddr{University of Washington}\\
%    \email{mcorrell@cs.washington.edu}}\\
%  \alignauthor{Jeffrey Heer\\
%    \affaddr{University of Washington}\\
%    \email{jheer@cs.washington.edu}}\\
}

\maketitle

\begin{abstract}
Observing trends in bivariate data, or predicting the future values of data, is a common task for viewers of visualizations. Yet, designs do not always explicitly draw trend lines and other relevant statistical modeling information in charts. Thus, viewers must often perform regression “by eye:” they estimate the trend of data through the values alone. Different design choices may aid or hamper this ability to estimate trends in data. In this work, we present a series of crowd-sourced experiments examining regression by eye, investigating both viewer performance at estimation of trends in bivariate data, and potential sources of bias in these estimations. Our findings indicate that viewers can accurately and robustly estimate trends in bivariate visualizations, but that certain features of both the data and the visual design of the visualization (such as outliers, asymmetry, and trend complexity) can negatively impact this accuracy.
\end{abstract}

\category{H.5.m.}{Information Interfaces and Presentation
  (e.g. HCI)}{Miscellaneous}

\keywords{\plainkeywords}

\section{Introduction}

\section{Acknowledgments}

Omitted for review.

% Balancing columns in a ref list is a bit of a pain because you
% either use a hack like flushend or balance, or manually insert
% a column break.  http://www.tex.ac.uk/cgi-bin/texfaq2html?label=balance
% multicols doesn't work because we're already in two-column mode,
% and flushend isn't awesome, so I choose balance.  See this
% for more info: http://cs.brown.edu/system/software/latex/doc/balance.pdf
%
% Note that in a perfect world balance wants to be in the first
% column of the last page.
%
% If balance doesn't work for you, you can remove that and
% hard-code a column break into the bbl file right before you
% submit:
%
% http://stackoverflow.com/questions/2149854/how-to-manually-equalize-columns-
% in-an-ieee-paper-if-using-bibtex
%
% Or, just remove \balance and give up on balancing the last page.
%
\balance{}

% REFERENCES FORMAT
% References must be the same font size as other body text.
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{sample}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2016-09-06 Paper/proceedings.tex
Paper outline

@@
 % so you remain anonymous.
-\def\plaintitle{Visual Estimation of Trend in Bivariate Visualizations}
+\def\plaintitle{Visual Estimation of Trend in Scatterplot Visualizations}
 \def\plainauthor{Removed for Review}
@@
 Observing trends in bivariate data, or predicting the future values of data, is a common task for viewers of visualizations.
-Yet, designs do not always explicitly draw trend lines and other relevant statistical modeling information in charts.
-Thus, viewers must often perform regression “by eye:” they estimate the trend of data through the values alone.
-Different design choices may aid or hamper this ability to estimate trends in data.
+Yet, designer do not always explicitly include trend lines and other relevant statistical modeling information in charts.
+Thus, viewers must often perform \emph{regression by eye} --- they estimate the trend of data through the values alone.
+If viewer performance at this task is sufficiently poor, or if certain designs introduce unacceptable bias, then designers will need to rethink how to communicate trends to the general audience.
 In this work, we present a series of crowd-sourced experiments examining regression by eye, investigating both viewer performance at estimation of trends in bivariate data, and potential sources of bias in these estimations.
-Our findings indicate that viewers can accurately and robustly estimate trends in bivariate visualizations, but that certain features of both the data and the visual design of the visualization (such as outliers, asymmetry, and trend complexity) can negatively impact this accuracy.
+Our findings indicate that, while viewers can accurately estimate trends in many standard visualizations of bivariate data, certain visual designs (such as the visual asymmetry of area charts) and certain features of the data (such as the presence of outliers) result in visual estimations that do not align with basic statistical models of regression.
 \end{abstract}
@@
 
+\section{Related Work}
+
+\section{General Methods}
+
+In order to assess the ability of viewers of visualizations to perform regression-by-eye, we conducted a series of three crowd-sourced experiments on Amazon's Mechanical Turk platform.
+We limited the participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
+For all experiments, we included stimuli where the trend line was visible as attention checks and for validation.
+We excluded participants with poor performance on these stimuli.
+Based on timings from internal piloting, we paid each participant \$2 for their participation, for an intended rate of \$8/hour.
+
+We generated the scatterplots by first generating the horizontal position of points through banded Monte Carlo sampling, ensuring that, while the distribution of points in X appeared uniform, at least X/b points appear in each of the b bands.
+We performed rejection sampling to ensure that no two points overlapped by more than half their radius in the resulting 600x1050 image.
+Next, we created a set of residual values, sampled evenly from discretized gaussian.
+This procedure ensured that the actual and intended bandwidth and standard deviation of residual values is precise.
+This set was permuted, and then applied to the points along with a target linear trend.
+Heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, and so, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
+We reused this residuals across all different trend types (linear, quadratic, or trigonometric).
+Except where noted, we selected trend lines that were centered in the image: that is, for a vertical and horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
+
+\subsection{Participants}
+
+\section{Estimation of Slope is Reliable \\ Across Visualizations}
+
+We presented participants with a series of scatterplots, and adjusted a slider to fit the perceived trend in the points.
+Scatterplots consisted of one of three types of trend (linear, quadratic, or trigonometric).
+For each scatterplot, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).
+
+\subsection{Hypotheses}
+\subsection{Results}
+\subsection{Discussion}
+
+\section{Estimation of Intercept is Subject to \\ ``Within The Bar'' Bias}
+\subsection{Hypotheses}
+\subsection{Results}
+\subsection{Discussion}
+
+\section{Estimation of Slope is Sensitive to Outliers}
+\subsection{Hypotheses}
+\subsection{Results}
+\subsection{Discussion}
+
+\section{Discussion}
+\subsection{Limitations \& Future Work}
+\subsection{Conclusion}
 \section{Acknowledgments}

### 2016-09-09 Paper/proceedings.tex
Paper Outline

@@
 
+One of the motivating examples for the power of information visualization is Anscombe's quartet\cite{anscombe1973graphs}, a set of four bivariate sets with nearly identical summary statistics, but with drastically different patterns when drawn as a scatterplot.
+Implicit in this example is that people have a robust ability to recognize visual patterns in information visualizations and, moreover, to employ this pattern recognition ability towards statistical ends.
+That is, an underlying assumption in information visualization is that people have the ability to \emph{visually} estimate \emph{statistical} quantities of interest, and can perform at least rough analogs of statistical tasks including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection.
+Knowing the limits of this ``estimation by eye,'' is important for designers of information visualizations seeking to communicate statistical quantities, especially to the general audience.
+
+In this work, we examine
+
 \section{Related Work}

### 2016-09-13 Paper/proceedings.tex
Skeleton version for paper swap.

@@
-\documentclass{sigchi}
+\documentclass[chi_draft]{sigchi}
+%\documentclass{sigchi}
 
@@
 
+%motivation here is poor. needs another turn of the crank.
+
+\todo{This motivation needs work. Also some redundancies}
+
 One of the motivating examples for the power of information visualization is Anscombe's quartet\cite{anscombe1973graphs}, a set of four bivariate sets with nearly identical summary statistics, but with drastically different patterns when drawn as a scatterplot.
@@
 Knowing the limits of this ``estimation by eye,'' is important for designers of information visualizations seeking to communicate statistical quantities, especially to the general audience.
+More than a mere mismatch between the statistical values which the designer intends to communicate, and the ability of viewers to estimate those values, there is also the possibility of \emph{bias}--- the design of visualizations may cause viewers to over- or underestimate task-relevant values.
 
-In this work, we examine
+The estimation of trends in bivariate data is an important analytical task, as it is the basis for many factors relevant to decision-making, such as prediction, imputation, and comparison.
+Unfortunately, model information is not always explicitly provided by designers of visualizations.
+When designers do include trend information (for instance, by annotating a scatterplot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent.
+Even if designers specifically include all relevant information, the audience may lack the statistical expertise to interpret these values.
+As a further complication, visual presentation of bivariate data can influence the types of analysis made by viewers (for instance, viewers may be more likely to consider trends with line charts, and to compare individual values with bar charts \cite{zacks1999bars}).
+These design choices can also introduce bias (for instance, the visual asymmetry of bars can cause ``within-the-bar'' bias \cite{newman2012bar}).
+Designers need guidance both on how much accuracy they can expect from viewers making trend estimations ``by eye,'' as well as whether or not different types of bivariate visualizations can bias these estimations.
+
+In this work, we describe a series of crowd-sourced studies on the visual estimation of trend in common bivariate visualizations (scatterplots, area charts, and linegraphs).
+We report on the results of three studies, investigating estimation of trend slope, trend intercept, and the effect of outliers.
+We find that, while in most cases viewers make accurate estimations of trends, area charts introduce systematic underestimation of trend intercept, and outliers introduce ambiguity in how trends are estimated.
+These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.
 
@@
 
+While there is a great deal of foundational work in visualization and graphical perception dealing with the estimation of individual values in visualizations (such as the height of the bar in a bar chart, or the angle of a line in a linegraph), there is comparatively little work on how viewers of visualizations perceive aggregate statistical quantities.
+Ariely~\cite{ariely2001seeing} suggests that, in concert with the perception of individual objects, we also collect information about the \emph{ensemble} properties of visual displays.
+Szafir et al.~\cite{szafir2016four} suggest that this \emph{ensemble coding} can afford the accurate estimation of \emph{summary statistics} in visualizations.
+However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}.
+
+Scatterplots are a standard design for visualization bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}.
+Prior work has confirmed that viewers can make use of scatterplots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}.
+However, the heuristics used when performing these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases.
+Similarly, the design of scatterplots (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}.
+The aspect ratio of graphs can also bias judgments about trend~\cite{beattie2002impact}--- wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts.
+Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.
+
+Recent work in the visualization community has focused on the perception of correlation in scatterplots.
+Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatterplots.
+Harrison et al.
+\cite{harrison2014ranking} extend this finding to other visualization types, although a reanalysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor.
+Estimation of correlation can also be biased by the design of scales~\cite{cleveland1982variables}--- the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.
+
+Our task of trend estimation combines elements of both prediction and correlation estimation tasks.
+As with prediction, there is not necessarily an unambiguous ``correct'' estimation of trend (different modeling and regression methods can produce different trend lines).
+As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.) likely relies on a set of key visual proxies.
+And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.
+
 \section{General Methods}
 
-In order to assess the ability of viewers of visualizations to perform regression-by-eye, we conducted a series of three crowd-sourced experiments on Amazon's Mechanical Turk platform.
+In order to assess the ability of viewers of visualizations to estimate trend, we conducted a series of three crowd-sourced experiments on Amazon's Mechanical Turk platform.
+Crowd-sourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}.
+Crowd-sourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{talbot2014four}.
+
 We limited the participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
@@
 
-We generated the scatterplots by first generating the horizontal position of points through banded Monte Carlo sampling, ensuring that, while the distribution of points in X appeared uniform, at least X/b points appear in each of the b bands.
+We generated the scatterplots by first generating the horizontal position of points through banded Monte Carlo sampling, ensuring that, while the distribution of points in $X$ appeared uniform, at least $X/b$ points appear in each of the $b$ bands.
 We performed rejection sampling to ensure that no two points overlapped by more than half their radius in the resulting 600x1050 image.
@@
 This set was permuted, and then applied to the points along with a target linear trend.
-Heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, and so, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
+As heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
 We reused this residuals across all different trend types (linear, quadratic, or trigonometric).
-Except where noted, we selected trend lines that were centered in the image: that is, for a vertical and horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
+Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
+
+\todo{Add figure of experimental apparatus}
+
+In all experiments, participants entered in information using a slider without tick marks, in order to avoid anchoring effects \cite{matejka2016effect}.
+Moving the slider adjusts a purple trend line by adjusting its slope (in experiments 1 and 3) or its y-intercept (in experiment 2).
+After moving the slider, participants had to confirm their choice of trend line.
+For validation, certain trials contained the actual (OLS) trend line.
+We excluded the results of participants with poor performance on these validation questions.
 
@@
 
-\section{Estimation of Slope is Reliable \\ Across Visualizations}
+\todo{48 per experiment, report all demographics here rather than in experiment subsections.}
 
-We presented participants with a series of scatterplots, and adjusted a slider to fit the perceived trend in the points.
-Scatterplots consisted of one of three types of trend (linear, quadratic, or trigonometric).
-For each scatterplot, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).
+\section{1. Estimation of Slope is Reliable \\ Across Visualizations}
+
+We presented participants with a series of bivariate visualizations, and adjusted a slider to fit the perceived trend in the points.
+Visualizations were one of a scatterplot, linegraph, or area chart (with the filled area below the line).
+Visualizations consisted of one of three types of trend (linear, quadratic, or trigonometric).
+For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).
 
@@
 
-\section{Estimation of Intercept is Subject to \\ ``Within The Bar'' Bias}
+\section{2. Estimation of Intercept is Subject to \\ ``Within The Bar'' Bias}
+
+As with the previous experiment, we presented participants with a series of bivariate visualizations.
+However, instead of estimating the \emph{slope} of the points, participants estimated the \emph{y intercept} of the trend line.
+For each trial, we added a uniform offset to the points in the bivariate visualization in the data range $[-0.25,0.25]$.
+The rendered trend line was initially placed with the correct slope, and such that $f(0.5)=0.5$ in data space.
+Participants adjusted a slider controlling the vertical offset of this trend line.
+
 \subsection{Hypotheses}
@@
 
-\section{Estimation of Slope is Sensitive to Outliers}
+\section{3. Estimation of Slope is Sensitive to Outliers}
+
+This experiment was largely identical to experiment 1 above, except we designated 0, 5, 10, or 15 points at the very beginning, middle, or end of the series as outliers, and placed them randomly in the top or bottom 10\% of the visual area of the visualization (whichever was furthest from the trend line).
+We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, intended trend line.
+As with experiment 1, the participants controlled the slope of a rendered trend line with a slider.
+However, rather than being placed such that $f(0.5)=0.5$, we offset the trend line to the intersection of the robust (without outliers) and non-robust (including outliers) trend line.
+This allowed participants to express both types of fit with the same slider interaction, while affording estimations beyond a simple interpolation of both trends.
+
 \subsection{Hypotheses}

### 2016-09-19 Paper/proceedings.tex
Final outlining, incorporating JH edits.

@@
-\documentclass[chi_draft]{sigchi}
-%\documentclass{sigchi}
+\documentclass{sigchi}
 
@@
 % so you remain anonymous.
-\def\plaintitle{Visual Estimation of Trend in Scatterplot Visualizations}
+\def\plaintitle{Regression by Eye: Estimating Trends in Bivariate Visualizations}
 \def\plainauthor{Removed for Review}
@@
 \begin{abstract}
-Observing trends in bivariate data, or predicting the future values of data, is a common task for viewers of visualizations.
-Yet, designer do not always explicitly include trend lines and other relevant statistical modeling information in charts.
-Thus, viewers must often perform \emph{regression by eye} --- they estimate the trend of data through the values alone.
-If viewer performance at this task is sufficiently poor, or if certain designs introduce unacceptable bias, then designers will need to rethink how to communicate trends to the general audience.
-In this work, we present a series of crowd-sourced experiments examining regression by eye, investigating both viewer performance at estimation of trends in bivariate data, and potential sources of bias in these estimations.
-Our findings indicate that, while viewers can accurately estimate trends in many standard visualizations of bivariate data, certain visual designs (such as the visual asymmetry of area charts) and certain features of the data (such as the presence of outliers) result in visual estimations that do not align with basic statistical models of regression.
+Observing trends and predicting future values are common tasks for viewers of bivariate data visualizations.
+As many charts do not explicitly include trend lines or related statistical summaries, viewers often visually estimate trends directly from a plot.
+How reliable are the inferences viewers draw when performing such \emph{regression by eye}?
+Do particular visualization designs or data features bias trend perception?
+We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations.
+We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models.
+In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge.
+These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
 \end{abstract}
@@
 
-\todo{This motivation needs work. Also some redundancies}
-
 One of the motivating examples for the power of information visualization is Anscombe's quartet\cite{anscombe1973graphs}, a set of four bivariate sets with nearly identical summary statistics, but with drastically different patterns when drawn as a scatterplot.
@@
 Knowing the limits of this ``estimation by eye,'' is important for designers of information visualizations seeking to communicate statistical quantities, especially to the general audience.
-More than a mere mismatch between the statistical values which the designer intends to communicate, and the ability of viewers to estimate those values, there is also the possibility of \emph{bias}--- the design of visualizations may cause viewers to over- or underestimate task-relevant values.
+There are two important classes of limitation: either the visual estimations are \emph{inaccurate} for the use cases intended by the designer, or they are \emph{biased}, creating systematic over- or under-estimations.
 
@@
 When designers do include trend information (for instance, by annotating a scatterplot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent.
+Viewers must therefore perform visual estimation to recover the relevant statistics not provided by the designers.
+
 Even if designers specifically include all relevant information, the audience may lack the statistical expertise to interpret these values.
@@
 These design choices can also introduce bias (for instance, the visual asymmetry of bars can cause ``within-the-bar'' bias \cite{newman2012bar}).
-Designers need guidance both on how much accuracy they can expect from viewers making trend estimations ``by eye,'' as well as whether or not different types of bivariate visualizations can bias these estimations.
+Designers need guidance both on how much accuracy they can expect from viewers making trend estimations by eye, as well as whether or not different types of bivariate visualizations can bias these estimations.
 
-In this work, we describe a series of crowd-sourced studies on the visual estimation of trend in common bivariate visualizations (scatterplots, area charts, and linegraphs).
+In this work, we describe a series of crowdsourced studies on the visual estimation of trends in common bivariate visualizations (scatterplots, area charts, and linegraphs).
 We report on the results of three studies, investigating estimation of trend slope, trend intercept, and the effect of outliers.
@@
 However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}.
+A further difficulty is that tasks requiring estimations of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can results in a multitude of potential valid responses).
 
@@
 Our task of trend estimation combines elements of both prediction and correlation estimation tasks.
-As with prediction, there is not necessarily an unambiguous ``correct'' estimation of trend (different modeling and regression methods can produce different trend lines).
+As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines).
 As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.) likely relies on a set of key visual proxies.
@@
 
-In order to assess the ability of viewers of visualizations to estimate trend, we conducted a series of three crowd-sourced experiments on Amazon's Mechanical Turk platform.
+In order to assess the ability of viewers of visualizations to estimate trend, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform.
+We designed these experiments to establish a baseline for performance at regression by eye, and to examine potential sources of bias for these estimates.
 Crowd-sourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}.
@@
 
-We limited the participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
-For all experiments, we included stimuli where the trend line was visible as attention checks and for validation.
-We excluded participants with poor performance on these stimuli.
-Based on timings from internal piloting, we paid each participant \$2 for their participation, for an intended rate of \$8/hour.
+In each experiment, we presented participants with a set of bivariate visualizations based on an underlying set of points.
+We wished to have precise and orthogonal control over relevant statistics such as noise, slope, and sample size, while maintaining the appearance of a ``natural'' distribution of points.
+Existing methods for generating points for related experimental tasks such as estimation of correlations (e.g., Harrison et al.~\cite{harrison2014ranking}) do not afford this orthogonality: e.g., Pearson's $r$ is correlated with the slope (points with identical residuals but different slopes of their linear fits will have differing $r$ values).
+We also wished to have a fair comparison between visual estimations and the results of OLS.
+We therefore used the standard model of OLS to generate points, namely that $y=\beta x + \epsilon$, where $\epsilon$ is an error term that follows a Gaussian model.
 
-We generated the scatterplots by first generating the horizontal position of points through banded Monte Carlo sampling, ensuring that, while the distribution of points in $X$ appeared uniform, at least $X/b$ points appear in each of the $b$ bands.
-We performed rejection sampling to ensure that no two points overlapped by more than half their radius in the resulting 600x1050 image.
-Next, we created a set of residual values, sampled evenly from discretized gaussian.
-This procedure ensured that the actual and intended bandwidth and standard deviation of residual values is precise.
+Therefore, for each point set, we created a set of residual values, sampled evenly from a discretized Gaussian.
 This set was permuted, and then applied to the points along with a target linear trend.
 As heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
-We reused this residuals across all different trend types (linear, quadratic, or trigonometric).
+We reused these residuals across all different trend types (linear, quadratic, or trigonometric).
 Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
-
-\todo{Add figure of experimental apparatus}
+In all experiments, we desired control over the direction of the trend.
+For linear fits, this is the slope of the trend line.
+For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term.
+For trignomoetric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes).
+These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.
 
@@
 After moving the slider, participants had to confirm their choice of trend line.
-For validation, certain trials contained the actual (OLS) trend line.
-We excluded the results of participants with poor performance on these validation questions.
+The central measure is accuracy, or the difference between the selected slope and the slope of the series if the residuals were removed.
+We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand.
+We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model-space of participants, and to reduce the impact of factors such as accuracy in motor movements and heterogeneity in input devices.
 
@@
 
-\todo{48 per experiment, report all demographics here rather than in experiment subsections.}
+We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
+For validation, certain trials contained the actual (OLS) trend line.
+We excluded the results of participants with poor performance on these validation questions.
+Across the three reported experiments and pilots, we performed 7 such exclusions.
+We recruited additional participants to replace these poor performers.
+Based on timings from internal piloting, we paid each participant \$2 for their participation, for an intended rate of \$8/hour.
+
+We analyzed 48 participants for each experiment (excluding rejections), for a total of 144 (98 male, 43 female, 3 who declined to stat, $M_{age}= 33.2$,$SD_{age}=8.8$) participants.
+The 10 had graduate or professional degrees, 69 had college degrees, 40 had at least some college, and the remaining 25 had high school diplomas.
+After completing the main experimental task, participants were asked to self-assess their familiarity with graphs and charts, on a 5 point Likert scale.
+The plurality (64) rated themselves as ``3.
+Some familiarity,'' and none rated themselves with the maximum rating of ``5.
+A great deal of experience.''
 
 \section{1. Estimation of Slope is Reliable \\ Across Visualizations}
+
+We designed our first experiment to examine how accurate participants were are estimating the direction (slope, amplitude, or curvature) of trends in bivariate visualizations.
+We examined a variety of common bivariate visualizations.
+In addition to linear trends, we examined more complex relationships such as quadratic and trignometric trends.
 
@@
 
+Participants saw one of each combination of 3 chart types (scatterplot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli.
+Type of trend (linear, quadratic, or trignometric) was a random factor, with 32 stimuli of each factor level.
+
 \subsection{Hypotheses}
+
+We had three hypotheses for the first experiment:
+\begin{enumerate}
+	\item \textbf{As the bandwidth of the residuals increased, accuracy would decrease}.
+Increasing the bandwidth of the residuals results in a lower correlation coefficient and higher perceived noise in the bivariate data.
+Prior work indicates that these related measures correspond to decreased accuracy for aggregate tasks in bivariate visualizations \cite{albers2014task,harrison2014ranking}.
+	\item \textbf{More complex relationships would result in lower accuracy}. 
+\end{enumerate}
 \subsection{Results}
-\subsection{Discussion}
 
@@
 
+``Within the bar'' bias is a perceptual bias in bar charts where points contained in the visual area of the glyph of the bar are seen as likelier than points outside of the glyph.
+Newman \& Scholl \cite{newman2012bar} encountered this bias for a sampling task: ``how likely is this point to have been drawn from the distribution represented by this bar?'' Correll \& Gleicher \cite{correll2014error} likewise encountered this bias for inferential tasks: ``how likely is the population mean to take a particular value, given the sample represented by this bar?'' We believed that this bias would likewise occur in regression tasks in visually asymmetric visualizations such as area charts.
+The slope estimation task from the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} underestimation in trends; decreasing the slope would cause underestimation at the beginning of the plot but not the end, and vice versa.
+We therefore designed this experiment to measure estimates of the y-intercepts of trends.
+The within-the-bar bias here would then appear as systematic underestimation of intercept in area charts.
+
 As with the previous experiment, we presented participants with a series of bivariate visualizations.
-However, instead of estimating the \emph{slope} of the points, participants estimated the \emph{y intercept} of the trend line.
+However, instead of estimating the \emph{slope} of the points, participants estimated the \emph{y-intercept} of the trend line.
 For each trial, we added a uniform offset to the points in the bivariate visualization in the data range $[-0.25,0.25]$.
@@
 
+The plots had the same factor levels as the previous experiment, for a total of 96 stimuli per participant. The uniform offset was an additional random factor for each stimulus.
+
 \subsection{Hypotheses}
+
+We had one hypothesis for the second experiment:
+\begin{enumerate}
+	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatterplots. 
+\end{enumerate}
 \subsection{Results}
-\subsection{Discussion}
 
@@
 
-This experiment was largely identical to experiment 1 above, except we designated 0, 5, 10, or 15 points at the very beginning, middle, or end of the series as outliers, and placed them randomly in the top or bottom 10\% of the visual area of the visualization (whichever was furthest from the trend line).
-We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, intended trend line.
+OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit.
+Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods.
+Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}.
+Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends.
+However, cognitive biases such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in their estimations.
+These biases can impact forecasting tasks~\cite{campbell2009anchoring,ji2001culture}.
+That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.
+
+This experiment was largely identical to experiment 1 above, except we designated 0, 5, 10, or 15 points at the very beginning, first third, or end of the series as outliers, and placed them randomly in the top or bottom 10\% of the visual area of the visualization (whichever was farthest from the trend line).
+We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, outlier-less trend line.
 As with experiment 1, the participants controlled the slope of a rendered trend line with a slider.
@@
 This allowed participants to express both types of fit with the same slider interaction, while affording estimations beyond a simple interpolation of both trends.
+This expressiveness is only afforded by linear fits of, and so we did not consider quadratic or trigonometric trends in this experiment.
+
+Similar to the previous experiments, participants saw one of each of the three graph types, slope, and Gaussian residual bandwidth.
+The four outlier quantities $\{0,5,10,15\}$ was an additional factor.
+To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively.
+This resulted in 96 total stimuli, in line with prior experiments.
 
 \subsection{Hypotheses}
+
+We had three hypotheses for the third experiment:
+\begin{enumerate}
+	\item \textbf{Participant estimation would be closer to an outlier-less robust trend line, than the non-robust OLS trend line.} That is, we assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
+	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
+	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in forecasting tasks suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'').
+We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
+\end{enumerate}
 \subsection{Results}
-\subsection{Discussion}
 

### 2016-09-20 Paper/figures/figures.tex
Figureless version.

\newcommand{\outlierFig}{
 \begin{figure}
 \centering
 \includegraphics[width=.55\columnwidth]{figures/exp4}
 \caption{
 	A cropped example stimulus from experiment 3. We replaced the final 15 points of this dataset with extreme values. The overlaid purple trend line represents a robust OLS fit (ignoring the outlier values), while the overlaid green line represents the fit with all points included. Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.
 }
 \label{fig:outlier}
 \end{figure}
}

\newcommand{\expFig}{
	\begin{figure}
		\centering
		\includegraphics[width=.65\columnwidth]{figures/exp}
		\caption{
			An example estimation task from our experiments. Here, the participant must adjust the amplitude of the purple trigonometric function until it best matches a particular set of bivariate data.
		}
		\label{fig:experiment}
	\end{figure}
}

### 2016-09-20 Paper/proceedings.tex
Figureless version.

@@
 
+\input{./figures/figures.tex}
+
 \section{Introduction}
@@
 
+\expFig
+
 In order to assess the ability of viewers of visualizations to estimate trend, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform.
@@
 Moving the slider adjusts a purple trend line by adjusting its slope (in experiments 1 and 3) or its y-intercept (in experiment 2).
+Fig.
+\ref{fig:experiment} shows an example experimental task.
 After moving the slider, participants had to confirm their choice of trend line.
@@
 We examined a variety of common bivariate visualizations.
-In addition to linear trends, we examined more complex relationships such as quadratic and trignometric trends.
+In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric trends.
 
@@
 
-Participants saw one of each combination of 3 chart types (scatterplot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli.
-Type of trend (linear, quadratic, or trignometric) was a random factor, with 32 stimuli of each factor level.
+Participants saw one of each combination of 3 chart types (scatterplot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli, with an additional 4 validation stimuli otherwise excluded from analysis.
+Type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.
 
@@
 	\item \textbf{More complex relationships would result in lower accuracy}.
+Quadratic and trigonometric relationships are visually more complex than linear relationships, and often require more complex statistical methods to analyze.
+We anticipated that estimation of these less familiar relationships would therefore be more difficult than in the linear case.
+	\item \textbf{Estimations would be unbiased}. That is, there would be no systematic over- or underestimation of trends. 
 \end{enumerate}
 \subsection{Results}
+We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on unsigned error in estimation of trend lines.
+We included participant ID and the actual slope of the trend line as covariates.
+
+Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}.
+Bandwidth of the Gaussian used to generate residuals was a significant main effect ($F(1,4554)=950$), and the trimmed mean of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$.
+Figure \ref{fig:sigma} illustrates these results.
+
+Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}.
+Fit type was only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc Tukey's test of Honest Significant Difference (HSD) did not identify any significant pairwise-interactions.
+The trimmed mean of absolute estimation error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a significant (in the ordinary sense) impact on performance.
+
+Our results support our third hypothesis: \textbf{there was no statistically significant bias in estimations}.
+Participants saw an equal number of stimuli with positive and negative trend lines.
+If estimates of these trends were unbiased, we would expect the average signed error to be close to zero.
+The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses (which had $\Delta=0.01$).
+A Student's T test failed to support ($p=0.70$) the hypothesis that $mu_{error}\ne0$.
 
@@
 
-The plots had the same factor levels as the previous experiment, for a total of 96 stimuli per participant. The uniform offset was an additional random factor for each stimulus.
+The plots had the same factor levels as the previous experiment, for a total of 96 stimuli per participant, with an additional 4 validation stimuli otherwise excluded from analysis.
+The uniform offset was an additional random factor for each stimulus.
 
@@
 \subsection{Results}
+We performed a one-way ANCOVA of the effect of graph type on \emph{signed} error. We included residual bandwidth as a covariate, and participant ID as a random factor.
 
-\section{3. Estimation of Slope is Sensitive to Outliers}
+Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatterplots or linegraphs}.
+Graph type was a significant main effect on signed error ($F(2,2431)=27$, $p<0.001$).
+A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatterplots and linegraphs.
+The trimmed mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions.
+This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for linegraphs and scatterplots).
+Figure \ref{fig:outlier} illustrates this result.
 
+\section{3. Estimation of Slope is Robust to Outliers}
+
+\outlierFig
 OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit.
@@
 To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively.
-This resulted in 96 total stimuli, in line with prior experiments.
+This resulted in 96 total stimuli, with an additional 4 validation stimuli otherwise excluded from analysis, in line with prior experiments.
 
@@
 
+Our results support our first hypothesis.
+\textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}.
+Excluding conditions with zero outliers (and so the two trend lines would be identical), the trimmed mean of unsigned error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$).
+A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($p<0.001$).
+However, a Student's T test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($M=0.07$ vs.
+$M=0.08$, $p<0.001$).
+This indicates that participants are not entirely ignoring outliers, although the small effect size suggestes that participants are giving outliers significantly less weight than OLS regression.
+
+%There is a confound between the number of outliers and the OLS trend line: since all outliers had approximately the same magnitude, more outliers create more extreme slopes with respect to the robust trend line, and so (if a participant were strictly ignoring outliers), larger differences in unsigned error.
+Similarly, outliers at the ends of the data series will adjust the slope of the OLS line more than outliers in the middle.
+We therefore measured performance as the unsigned error as a percentage of the difference between the slopes of the two types of fit.
+We performed a two-way ANOVA on the effect of outlier number and location on this percentage error measure.
+Participant id was included as a random variable.
+
+%Our results fail to support our second hypothesis.
+\emph{Increasing number of outliers does not increase fidelity to the OLS trend line}.
+A one-way ANCOVA on the effect of outlier numbers greater than 0 on this percentage error showed that outliers were a significant main effect ($F(1,3438)=59$, $p<0.001$).
+However, the percentage error \emph{increases} as the number of outliers increase.
+A post-hoc Tukey's HSD shows that the percentage errors for 10 and 15 outliers are similar (trimmed means are $111\%$ and $112\%$ respectively), they are significantly different from the stimuli with only 5 outliers (trimmed mean of $97\%$).
+
 \section{Discussion}
+
+In many cases, designers do not explicitly encoded regression information.
+In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied.
+Yet, our results point to regression by eye as a robust and accurate method for estimating trends in bivariate data.
+Participants from a variety of backgrounds and levels of self-reported graphical and statistical expertise were capable of estimating both the slope and intercept of both linear and non-linear trends.
+That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.
+
+However, this general accuracy at estimation trends is not universal.
+Area charts are visually asymmetrical: area below the line is filled in, and the area above it as not.
+This asymmetry creates within the bar bias: a systematic underestimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area.
+Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.
+
+Likewise, viewers do not give the same weight to outliers as certain standard forms of statistical regression such as OLS.
+For noisy data, this robustness may be beneficial: humans themselves can be relied on to perform filtering operations without explicit guidance.
+However, in other cases, this insensitivity may result in visual estimations at odds with the statistical modeling underpinning other areas of analysis of a particular dataset.
+The emergence of new modes and the shifting of means may not result in a subsequent shift in beliefs about trends in data.
+
 \subsection{Limitations \& Future Work}
+
+Our experimental setup was intentionally simplistic, affording only a single free parameter for each experiment.
+In actual regression by eye, the viewer must simultaneously engage in multiple types of estimation: choosing a particular type of fit, filtering data, and then estimating the parameters of the chosen model.
+Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are presented as a given.
+
+There are also a number of design decisions not considered in this study that could impact regression by eye.
+For example, Wood et al.~\cite{wood2012sketchy} have shown that ``sketchiness'' in visualizations can result in skepticism of the data and design.
+It is possible that outliers could receive less visual (and so statistical) weight as a response to this sort of skepticism.
+Our stimuli likewise contained a sufficient number of points that bar charts were not a feasible choice of visualization.
+Given the propensity of bar charts to encourage comparison of individual, rather than aggregate quantities~\cite{zacks1999bars}, it is possible that bar charts of smaller scale bivariate data could promote fitting of local rather than global trends.
+
+Finally, we focused on a simple form of regression, OLS, as our gold standard for measuring accuracy.
+While our data were constructed to satisfy the assumptions of OLS (with the exception of our outlier experiment), in most real world scenarios OLS is just one tool of many, and analysts must exercise judgment when determining how to fit their data.
+More complex forms of regression, and more complex fits, may not have ready analogues visually, and data concerning regression by eye may not extend to cover these cases.
+
+Our future work is focused in three areas.
+Firstly, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye.
+Can sufficient information promote caution in judgments of trend, or do initial visual estimations have too much weight to be overcome?
+Secondly, we wish to examine the impact of different rhetorical framings and presentations on regression by eye.
+Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings (and so associated predispositions or biases about the data) could similarly impact statistical judgments.
+Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities.
+Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization \cite{micallef2012assessing}.
+
 \subsection{Conclusion}
+
+In this paper, we examine the ability of viewers of visualizations to perform regression by eye --- the visual estimation of trend in bivariate visualizations.
+We show that even viewers without statistical training can reliably estimate even non-linear trends in visualizations like scatterplots and linegraphs.
+However, area charts are subject to within the bar bias, and promote trends with lower intercepts than other bivariate visualizations.
+Regression by eye is also less sensitive to outliers than simple statistical regression techniques, resulting in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, can give equal weight all points in a dataset.
+
 \section{Acknowledgments}

### 2016-09-20 Paper/proceedings.tex
Update intro, add gitignore.

@@
 \usepackage{todonotes}
+
+\newcommand{\jeff}[1]{\textcolor{red}{JMH: #1}}
 
@@
 These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
+\jeff{These last three sentences seem a bit overlapping / redundant. Plus we really seem to like the word ``systematically''.}
 \end{abstract}
@@
 
-One of the motivating examples for the power of information visualization is Anscombe's quartet\cite{anscombe1973graphs}, a set of four bivariate sets with nearly identical summary statistics, but with drastically different patterns when drawn as a scatterplot.
-Implicit in this example is that people have a robust ability to recognize visual patterns in information visualizations and, moreover, to employ this pattern recognition ability towards statistical ends.
-That is, an underlying assumption in information visualization is that people have the ability to \emph{visually} estimate \emph{statistical} quantities of interest, and can perform at least rough analogs of statistical tasks including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection.
-Knowing the limits of this ``estimation by eye,'' is important for designers of information visualizations seeking to communicate statistical quantities, especially to the general audience.
-There are two important classes of limitation: either the visual estimations are \emph{inaccurate} for the use cases intended by the designer, or they are \emph{biased}, creating systematic over- or under-estimations.
+An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}, a set of four bivariate datasets with nearly identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots.
+\jeff{Add figure of Anscombe's quartet...} This example relies on the fact that people have the ability to \emph{perceptually} estimate statistical quantities of interest.
+Visualization users regularly perform statistical tasks, including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection.
+\jeff{Citations to perceptual work touching on the previous points?
+E.g., scatter plot means, color weaving, others?}
 
 The estimation of trends in bivariate data is an important analytical task, as it is the basis for many factors relevant to decision-making, such as prediction, imputation, and comparison.
-Unfortunately, model information is not always explicitly provided by designers of visualizations.
+However, model information is not always explicitly included by visualization designers.
 When designers do include trend information (for instance, by annotating a scatterplot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent.
-Viewers must therefore perform visual estimation to recover the relevant statistics not provided by the designers.
+Viewers must therefore perform visual estimation to gain a sense of any relevant statistics not provided.
 
-Even if designers specifically include all relevant information, the audience may lack the statistical expertise to interpret these values.
-As a further complication, visual presentation of bivariate data can influence the types of analysis made by viewers (for instance, viewers may be more likely to consider trends with line charts, and to compare individual values with bar charts \cite{zacks1999bars}).
-These design choices can also introduce bias (for instance, the visual asymmetry of bars can cause ``within-the-bar'' bias \cite{newman2012bar}).
-Designers need guidance both on how much accuracy they can expect from viewers making trend estimations by eye, as well as whether or not different types of bivariate visualizations can bias these estimations.
+Even if designers include all relevant information, the audience may lack the statistical expertise to interpret these values, or may be misled if the data violate modeling assumptions.
+As a further complication, the form of visual encoding may influence viewers' inferences.
+For instance, viewers may be more likely to consider trends with line charts, and to compare individual values with bar charts \cite{zacks1999bars}.
+Visual design choices can also introduce bias, such as the visual asymmetry of bars causing ``within-the-bar'' bias \cite{newman2012bar}.
+Designers would benefit from guidance regarding how accurately viewers make trend estimations by eye, and to what degree different visualization types and annotations might bias these estimations.
 
-In this work, we describe a series of crowdsourced studies on the visual estimation of trends in common bivariate visualizations (scatterplots, area charts, and linegraphs).
-We report on the results of three studies, investigating estimation of trend slope, trend intercept, and the effect of outliers.
-We find that, while in most cases viewers make accurate estimations of trends, area charts introduce systematic underestimation of trend intercept, and outliers introduce ambiguity in how trends are estimated.
-These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.
+Knowing the strengths and limitations of such ``estimation by eye'' is therefore important for designers of data visualizations seeking to communicate statistical quantities, especially to a general audience.
+On the one hand, visual estimations may be \emph{inaccurate} for the use cases intended by a designer, or they may be \emph{biased}, leading to systematic over- or under-estimations.
+On the other hand, while visual estimation lacks the precision of formal statistics, it may be relatively unencumbered by modeling assumptions.
+
+In this work, we describe a series of crowdsourced experiments on the visual estimation of trends in common bivariate visualizations such as scatter plots, area charts, and line graphs.
+We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers.
+We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and outliers introduce ambiguity in how trends are estimated.
+\jeff{This is itself a bit ambiguous.
+What do you mean by ``introduce ambiguity''?
+Can you be more precise?} These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.
+\jeff{Obligatory design implications remark here?}
 

### 2016-09-20 Paper/proceedings.tex
Comments on related work.

@@
 
-While there is a great deal of foundational work in visualization and graphical perception dealing with the estimation of individual values in visualizations (such as the height of the bar in a bar chart, or the angle of a line in a linegraph), there is comparatively little work on how viewers of visualizations perceive aggregate statistical quantities.
+While there is a great deal of foundational work in visualization and graphical perception dealing with the estimation of individual values in visualizations (such as the height of the bar in a bar chart, or the angle of a line in a line graph), there is comparatively little work on how viewers of visualizations perceive aggregate statistical quantities.
+
 Ariely~\cite{ariely2001seeing} suggests that, in concert with the perception of individual objects, we also collect information about the \emph{ensemble} properties of visual displays.
-Szafir et al.~\cite{szafir2016four} suggest that this \emph{ensemble coding} can afford the accurate estimation of \emph{summary statistics} in visualizations.
+Szafir et al.~\cite{szafir2016four} note that this \emph{ensemble coding} might afford relatively accurate estimation of \emph{summary statistics} in visualizations.
 However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}.
-A further difficulty is that tasks requiring estimations of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can results in a multitude of potential valid responses).
+A further difficulty is that tasks requiring estimation of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can result in a multitude of potential valid responses).
 
-Scatterplots are a standard design for visualization bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}.
-Prior work has confirmed that viewers can make use of scatterplots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}.
+Scatter plots are a standard means of visualizing bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}.
+Prior work has confirmed that viewers can make use of scatter plots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}.
 However, the heuristics used when performing these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases.
-Similarly, the design of scatterplots (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}.
-The aspect ratio of graphs can also bias judgments about trend~\cite{beattie2002impact}--- wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts.
+
+Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}.
+The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}\,---\,wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts.
 Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.
 
-Recent work in the visualization community has focused on the perception of correlation in scatterplots.
-Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatterplots.
-Harrison et al.
-\cite{harrison2014ranking} extend this finding to other visualization types, although a reanalysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor.
-Estimation of correlation can also be biased by the design of scales~\cite{cleveland1982variables}--- the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.
+Recent work in the visualization community has focused on the perception of correlation in scatter plots.
+Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatter plots.
+Harrison et al.~\cite{harrison2014ranking} extend this finding to other visualization types, although a re-analysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor.
+Estimation of correlation can also be biased by the choice of scales~\cite{cleveland1982variables}\,---\,the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.
 
@@
 As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines).
-As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.) likely relies on a set of key visual proxies.
-And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.
+As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of key visual proxies.
+\jeff{What do you mean by ``key visual proxies''?
+Be clearer so that readers can follow along.} And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.
+\jeff{Such as? The related work could end on a stronger note if we put forward some concrete hypotheses and note that prior work has not addressed them...}
 

### 2016-09-20 Paper/figures/figures.tex
Updates to general methods.

@@
   \includegraphics[width=.55\columnwidth]{figures/exp4}
- \caption{
- 	A cropped example stimulus from experiment 3.
+  \caption{A cropped example stimulus from experiment 3.
 We replaced the final 15 points of this dataset with extreme values.
 The overlaid purple trend line represents a robust OLS fit (ignoring the outlier values), while the overlaid green line represents the fit with all points included.
-Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.
- }
+Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.}
   \label{fig:outlier}
@@
   \includegraphics[width=.65\columnwidth]{figures/exp}
-		\caption{
-			An example estimation task from our experiments. Here, the participant must adjust the amplitude of the purple trigonometric function until it best matches a particular set of bivariate data.
-		}
+  \caption{An example estimation task from our experiments.
+Here, the participant must adjust the amplitude of the purple trigonometric function until it best matches a particular set of bivariate data.}
   \label{fig:experiment}

### 2016-09-20 Paper/proceedings.tex
Updates to general methods.

@@
 However, model information is not always explicitly included by visualization designers.
-When designers do include trend information (for instance, by annotating a scatterplot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent.
+When designers do include trend information (for instance, by annotating a scatter plot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent.
 Viewers must therefore perform visual estimation to gain a sense of any relevant statistics not provided.
@@
 
-In order to assess the ability of viewers of visualizations to estimate trend, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform.
-We designed these experiments to establish a baseline for performance at regression by eye, and to examine potential sources of bias for these estimates.
-Crowd-sourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}.
-Crowd-sourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{talbot2014four}.
-
-In each experiment, we presented participants with a set of bivariate visualizations based on an underlying set of points.
-We wished to have precise and orthogonal control over relevant statistics such as noise, slope, and sample size, while maintaining the appearance of a ``natural'' distribution of points.
-Existing methods for generating points for related experimental tasks such as estimation of correlations (e.g., Harrison et al.~\cite{harrison2014ranking}) do not afford this orthogonality: e.g., Pearson's $r$ is correlated with the slope (points with identical residuals but different slopes of their linear fits will have differing $r$ values).
-We also wished to have a fair comparison between visual estimations and the results of OLS.
-We therefore used the standard model of OLS to generate points, namely that $y=\beta x + \epsilon$, where $\epsilon$ is an error term that follows a Gaussian model.
-
-Therefore, for each point set, we created a set of residual values, sampled evenly from a discretized Gaussian.
-This set was permuted, and then applied to the points along with a target linear trend.
-As heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
-We reused these residuals across all different trend types (linear, quadratic, or trigonometric).
-Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
-In all experiments, we desired control over the direction of the trend.
-For linear fits, this is the slope of the trend line.
-For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term.
-For trignomoetric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes).
-These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.
-
-In all experiments, participants entered in information using a slider without tick marks, in order to avoid anchoring effects \cite{matejka2016effect}.
-Moving the slider adjusts a purple trend line by adjusting its slope (in experiments 1 and 3) or its y-intercept (in experiment 2).
-Fig.
-\ref{fig:experiment} shows an example experimental task.
-After moving the slider, participants had to confirm their choice of trend line.
-The central measure is accuracy, or the difference between the selected slope and the slope of the series if the residuals were removed.
-We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand.
-We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model-space of participants, and to reduce the impact of factors such as accuracy in motor movements and heterogeneity in input devices.
+In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform.
+We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias for these estimates.
+Crowdsourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}.
+Crowdsourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}.
+In this section, we describe experimental design aspects shared across all experiments.
 
@@
 Across the three reported experiments and pilots, we performed 7 such exclusions.
-We recruited additional participants to replace these poor performers.
-Based on timings from internal piloting, we paid each participant \$2 for their participation, for an intended rate of \$8/hour.
+We recruited additional participants to replace these excluded subjects.
+Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.
 
-We analyzed 48 participants for each experiment (excluding rejections), for a total of 144 (98 male, 43 female, 3 who declined to stat, $M_{age}= 33.2$,$SD_{age}=8.8$) participants.
-The 10 had graduate or professional degrees, 69 had college degrees, 40 had at least some college, and the remaining 25 had high school diplomas.
-After completing the main experimental task, participants were asked to self-assess their familiarity with graphs and charts, on a 5 point Likert scale.
+We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $\mu_{age}= 33.2$, $\sigma_{age}=8.8$).
+Across the subject pool, 10 reported having graduate or professional degrees, 69 college degrees, 40 at least some college, and 25 high school diplomas.
+After completing the main experimental task, participants were asked to self-assess their familiarity with charts and graphs on a 5 point Likert scale.
 The plurality (64) rated themselves as ``3.
@@
 A great deal of experience.''
+
+\subsection{Experimental Interface}
+
+In each experiment, we presented participants with visualizations of bivariate data and asked them to interactively adjust trend lines to best fit that data.
+Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}.
+Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2).
+Fig.
+\ref{fig:experiment} shows an example experimental task.
+After moving the slider, participants had to confirm their choice of trend line.
+The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed).
+We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand.
+We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model space and to reduce the impact of factors such as accuracy of motor movements or heterogeneity of input devices.
+
+\subsection{Data Generation}
+
+For each stimulus we wished to have precise, independent control over relevant statistics such as noise, slope, and sample size, while maintaining the appearance of a ``natural'' distribution of points.
+Existing methods for generating points for related experimental tasks such as estimation of correlations (e.g., Harrison et al.~\cite{harrison2014ranking}) do not afford independence: e.g., Pearson's $r$ is correlated with the slope, as points with identical residuals but different slopes of their linear fits will have differing $r$ values.
+We also wished to have a fair comparison between visual estimation and the results of ordinary least squares (OLS) regression.
+We therefore used the standard model of OLS to generate points, namely that $y=\beta x + \epsilon$, where $\epsilon$ is a normally-distributed error term.
+
+For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian.
+This set was permuted, and then applied to the points along with a target trend.
+As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
+We reused these residuals across all different trend types (linear, quadratic, or trigonometric).
+Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
+In all experiments, we desired control over the direction of the trend.
+For linear fits, this is the slope of the trend line.
+For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term.
+For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes).
+These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.
 
@@
 We presented participants with a series of bivariate visualizations, and adjusted a slider to fit the perceived trend in the points.
-Visualizations were one of a scatterplot, linegraph, or area chart (with the filled area below the line).
+Visualizations were one of a scatter plot, linegraph, or area chart (with the filled area below the line).
 Visualizations consisted of one of three types of trend (linear, quadratic, or trigonometric).
@@
 
-Participants saw one of each combination of 3 chart types (scatterplot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli, with an additional 4 validation stimuli otherwise excluded from analysis.
+Participants saw one of each combination of 3 chart types (scatter plot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli, with an additional 4 validation stimuli otherwise excluded from analysis.
 Type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.
@@
 \begin{enumerate}
-	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatterplots.
+	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatter plots.
 \end{enumerate}
@@
 
-Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatterplots or linegraphs}.
+Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatter plots or linegraphs}.
 Graph type was a significant main effect on signed error ($F(2,2431)=27$, $p<0.001$).
-A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatterplots and linegraphs.
+A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and linegraphs.
 The trimmed mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions.
-This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for linegraphs and scatterplots).
+This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for linegraphs and scatter plots).
 Figure \ref{fig:outlier} illustrates this result.
@@
 In this paper, we examine the ability of viewers of visualizations to perform regression by eye --- the visual estimation of trend in bivariate visualizations.
-We show that even viewers without statistical training can reliably estimate even non-linear trends in visualizations like scatterplots and linegraphs.
+We show that even viewers without statistical training can reliably estimate even non-linear trends in visualizations like scatter plots and linegraphs.
 However, area charts are subject to within the bar bias, and promote trends with lower intercepts than other bivariate visualizations.

### 2016-09-20 Paper/figures/figures.tex
Comments for remainder of draft.

@@
 The overlaid purple trend line represents a robust OLS fit (ignoring the outlier values), while the overlaid green line represents the fit with all points included.
-Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.}
+Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.
+\jeff{Could we include the average trend line produced by participants?}}
   \label{fig:outlier}

### 2016-09-20 Paper/proceedings.tex
Comments for remainder of draft.

@@
 Visual design choices can also introduce bias, such as the visual asymmetry of bars causing ``within-the-bar'' bias \cite{newman2012bar}.
-Designers would benefit from guidance regarding how accurately viewers make trend estimations by eye, and to what degree different visualization types and annotations might bias these estimations.
+Designers would benefit from guidance regarding how accurately viewers make trend estimations by eye, and to what degree different visualization types might bias these estimations.
 
@@
 
-\section{1. Estimation of Slope is Reliable \\ Across Visualizations}
+\section{Experiment 1: Slope Estimation}
 
-We designed our first experiment to examine how accurate participants were are estimating the direction (slope, amplitude, or curvature) of trends in bivariate visualizations.
-We examined a variety of common bivariate visualizations.
-In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric trends.
+We designed our first experiment to examine how accurate participants were at estimating the magnitude (slope, amplitude, or curvature) of trends in bivariate visualizations.
+We examined a variety of common bivariate visualizations \jeff{Such as?}.
+In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.
 
-We presented participants with a series of bivariate visualizations, and adjusted a slider to fit the perceived trend in the points.
-Visualizations were one of a scatter plot, linegraph, or area chart (with the filled area below the line).
-Visualizations consisted of one of three types of trend (linear, quadratic, or trigonometric).
+We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend in the points.
+The visualizations were one of a scatter plot, line graph, or area chart (with the filled area below the line).
+The slider parameterized one of three types of trend (linear, quadratic, or trigonometric).
 For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).
+\jeff{To many parentheticals...}
 
-Participants saw one of each combination of 3 chart types (scatter plot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli, with an additional 4 validation stimuli otherwise excluded from analysis.
-Type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.
+Participants saw one of each combination of 3 chart types (scatter plot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, and 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli.
+We also included an additional 4 validation stimuli otherwise excluded from analysis.
+The type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.
+\jeff{Was this factor random or consistent across subjects?}
 
@@
 Increasing the bandwidth of the residuals results in a lower correlation coefficient and higher perceived noise in the bivariate data.
-Prior work indicates that these related measures correspond to decreased accuracy for aggregate tasks in bivariate visualizations \cite{albers2014task,harrison2014ranking}.
+Prior work indicates that these related measures correspond to decreased accuracy for aggregate tasks in bivariate visualizations~\cite{albers2014task, harrison2014ranking}.
 	\item \textbf{More complex relationships would result in lower accuracy}.
 Quadratic and trigonometric relationships are visually more complex than linear relationships, and often require more complex statistical methods to analyze.
-We anticipated that estimation of these less familiar relationships would therefore be more difficult than in the linear case.
+We anticipated that estimation of these less familiar relationships would therefore be more difficult than the linear case.
 	\item \textbf{Estimations would be unbiased}. That is, there would be no systematic over- or underestimation of trends.
 \end{enumerate}
+
 \subsection{Results}
-We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on unsigned error in estimation of trend lines.
-We included participant ID and the actual slope of the trend line as covariates.
+We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on unsigned error \jeff{absolute error?} in estimation of trend lines.
+We included participant ids and the actual slope of the trend line as covariates.
 
 Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}.
-Bandwidth of the Gaussian used to generate residuals was a significant main effect ($F(1,4554)=950$), and the trimmed mean of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$.
+We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, \jeff{p-value?}).
+The trimmed mean \jeff{trimmed how?} of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$.
 Figure \ref{fig:sigma} illustrates these results.
@@
 Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}.
-Fit type was only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc Tukey's test of Honest Significant Difference (HSD) did not identify any significant pairwise-interactions.
-The trimmed mean of absolute estimation error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a significant (in the ordinary sense) impact on performance.
+Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise-interactions.
+The trimmed mean of absolute estimation error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.
 
@@
 If estimates of these trends were unbiased, we would expect the average signed error to be close to zero.
-The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses (which had $\Delta=0.01$).
-A Student's T test failed to support ($p=0.70$) the hypothesis that $mu_{error}\ne0$.
+The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses ($\Delta=0.01$).
+A Student's T-test failed to support the hypothesis that $mu_{error}\ne0$ ($p=0.70$) \jeff{Test statistic and degrees of freedom?}.
 
-\section{2. Estimation of Intercept is Subject to \\ ``Within The Bar'' Bias}
+\section{Experiment 2: ``Within-the-Bar'' Bias}
 
-``Within the bar'' bias is a perceptual bias in bar charts where points contained in the visual area of the glyph of the bar are seen as likelier than points outside of the glyph.
-Newman \& Scholl \cite{newman2012bar} encountered this bias for a sampling task: ``how likely is this point to have been drawn from the distribution represented by this bar?'' Correll \& Gleicher \cite{correll2014error} likewise encountered this bias for inferential tasks: ``how likely is the population mean to take a particular value, given the sample represented by this bar?'' We believed that this bias would likewise occur in regression tasks in visually asymmetric visualizations such as area charts.
-The slope estimation task from the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} underestimation in trends; decreasing the slope would cause underestimation at the beginning of the plot but not the end, and vice versa.
-We therefore designed this experiment to measure estimates of the y-intercepts of trends.
-The within-the-bar bias here would then appear as systematic underestimation of intercept in area charts.
+``Within the bar'' bias is a known perceptual bias involving bar charts, in which points contained in the visual area of the glyph of the bar are deemed likelier than points outside of the glyph.
+Newman \& Scholl~\cite{newman2012bar} encountered this bias for a sampling task: ``how likely is this point to have been drawn from the distribution represented by this bar?'' Correll \& Gleicher~\cite{correll2014error} likewise encountered this bias for inferential tasks: ``how likely is the population mean to take a particular value, given the sample represented by this bar?''
+
+We hypothesized that this bias would likewise occur for regression tasks when using visually asymmetric visualizations such as area charts.
+The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} underestimation in trends; decreasing the slope would cause underestimation at the beginning of the plot but not the end, and vice versa.
+We therefore designed this experiment to elicit estimates of the y-intercepts of trends.
+A within-the-bar bias would then appear as systematic underestimation of intercept in area charts.
 
@@
 \end{enumerate}
+
 \subsection{Results}
-We performed a one-way ANCOVA of the effect of graph type on \emph{signed} error. We included residual bandwidth as a covariate, and participant ID as a random factor.
+We performed a one-way ANCOVA of the effect of graph type on \emph{signed} error. We included residual bandwidth as a covariate, and participant ids as a random factor.
 
-Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatter plots or linegraphs}.
-Graph type was a significant main effect on signed error ($F(2,2431)=27$, $p<0.001$).
-A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and linegraphs.
-The trimmed mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions.
-This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for linegraphs and scatter plots).
+Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatter plots or line graphs}.
+We observed a significant main effect of graph type on signed error ($F(2,2431)=27$, $p<0.001$).
+A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and line graphs.
+The trimmed mean \jeff{trimmed how?} of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions.
+This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots).
 Figure \ref{fig:outlier} illustrates this result.
 
-\section{3. Estimation of Slope is Robust to Outliers}
+\section{Experiment 3. Estimation involving Outliers}
 
@@
 Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends.
-However, cognitive biases such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in their estimations.
-These biases can impact forecasting tasks~\cite{campbell2009anchoring,ji2001culture}.
+However, cognitive biases such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in their estimations \jeff{Need citations in this sentence?}.
+These biases can impact forecasting tasks~\cite{campbell2009anchoring, ji2001culture}.
 That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.
@@
 \begin{enumerate}
-	\item \textbf{Participant estimation would be closer to an outlier-less robust trend line, than the non-robust OLS trend line.} That is, we assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
+	\item \textbf{Participant estimation would be closer to a robust trend line ignoring outliers, than the non-robust OLS trend line.} That is, we assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
 	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
-	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in forecasting tasks suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'').
+	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in forecasting tasks suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'') \jeff{cite?}.
 We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
 \end{enumerate}
+
 \subsection{Results}
@@
 \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}.
-Excluding conditions with zero outliers (and so the two trend lines would be identical), the trimmed mean of unsigned error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$).
-A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($p<0.001$).
-However, a Student's T test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($M=0.07$ vs.
-$M=0.08$, $p<0.001$).
+Excluding conditions with zero outliers (and so the two trend lines would be identical), the trimmed mean \jeff{trimmed how?} of unsigned error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$).
+A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($p<0.001$) \jeff{test statistic, dof}.
+However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs.
+$\mu=0.08$, $p<0.001$).
 This indicates that participants are not entirely ignoring outliers, although the small effect size suggestes that participants are giving outliers significantly less weight than OLS regression.
+
+\jeff{Whither the other hypotheses?}
 
@@
 
-In many cases, designers do not explicitly encoded regression information.
+In many cases, designers do not explicitly encode regression information.
 In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied.
@@
 That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.
+\jeff{This point was not so clear in the results discussion of experiment 1. We should discuss (and visualize) the range of accuracies observed there.}
 
-However, this general accuracy at estimation trends is not universal.
-Area charts are visually asymmetrical: area below the line is filled in, and the area above it as not.
-This asymmetry creates within the bar bias: a systematic underestimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area.
+However, this general accuracy of trend estimation is not universal.
+Area charts are visually asymmetrical: the area below the line is filled in, and the area above it is not.
+We found that this asymmetry creates a within the bar bias: a systematic underestimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area.
 Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.
 
-Likewise, viewers do not give the same weight to outliers as certain standard forms of statistical regression such as OLS.
-For noisy data, this robustness may be beneficial: humans themselves can be relied on to perform filtering operations without explicit guidance.
-However, in other cases, this insensitivity may result in visual estimations at odds with the statistical modeling underpinning other areas of analysis of a particular dataset.
+Likewise, viewers do not give the same weight to outliers as OLS regression, but they do not ignore them either.
+For noisy data, this robustness may be beneficial: people can be relied on to perform filtering operations without explicit guidance.
+However, in other cases, this insensitivity may result in visual estimations at odds with the statistical modeling underpinning other areas of analysis of a particular dataset \jeff{What are you trying to say here?}.
 The emergence of new modes and the shifting of means may not result in a subsequent shift in beliefs about trends in data.
+\jeff{???}
 
@@
 
-Our experimental setup was intentionally simplistic, affording only a single free parameter for each experiment.
-In actual regression by eye, the viewer must simultaneously engage in multiple types of estimation: choosing a particular type of fit, filtering data, and then estimating the parameters of the chosen model.
-Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are presented as a given.
+Our experimental setup was intentionally simple, affording only a single free parameter for each experiment.
+In actual regression by eye, the viewer may simultaneously engage in multiple types of estimation: choosing a particular type of fit, ignoring outliers, and estimating the parameters of the chosen model.
+Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are fixed.
 
@@
 
-Finally, we focused on a simple form of regression, OLS, as our gold standard for measuring accuracy.
-While our data were constructed to satisfy the assumptions of OLS (with the exception of our outlier experiment), in most real world scenarios OLS is just one tool of many, and analysts must exercise judgment when determining how to fit their data.
-More complex forms of regression, and more complex fits, may not have ready analogues visually, and data concerning regression by eye may not extend to cover these cases.
+Finally, we focused on a simple form of regression, OLS, as our standard for measuring accuracy.
+While our data were constructed to satisfy the assumptions of OLS (with the intentional exception of our outlier experiment), in most real world scenarios OLS is just one tool of many, and analysts must exercise judgment when determining how to fit their data.
+More complex models may not have ready visual analogues, and data concerning regression by eye may not extend to cover these cases.
 
 Our future work is focused in three areas.
-Firstly, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye.
+First, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye.
 Can sufficient information promote caution in judgments of trend, or do initial visual estimations have too much weight to be overcome?
-Secondly, we wish to examine the impact of different rhetorical framings and presentations on regression by eye.
+Second, we wish to examine the impact of different rhetorical framings and presentations on regression by eye.
 Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings (and so associated predispositions or biases about the data) could similarly impact statistical judgments.
-Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities.
-Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization \cite{micallef2012assessing}.
+\jeff{Can you clarify this last part?
+And/or provide a concrete example?} Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities.
+Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization~\cite{micallef2012assessing}.
 
@@
 
-In this paper, we examine the ability of viewers of visualizations to perform regression by eye --- the visual estimation of trend in bivariate visualizations.
-We show that even viewers without statistical training can reliably estimate even non-linear trends in visualizations like scatter plots and linegraphs.
-However, area charts are subject to within the bar bias, and promote trends with lower intercepts than other bivariate visualizations.
+In this paper, we examine the ability of viewers of visualizations to perform \emph{regression by eye}\,---\,the visual estimation of trends in bivariate visualizations.
+We show that even viewers without statistical training can reliably estimate both linear and non-linear trends in visualizations such as scatter plots and line graphs.
+However, area charts are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations.
 Regression by eye is also less sensitive to outliers than simple statistical regression techniques, resulting in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, can give equal weight all points in a dataset.
+\jeff{Run-on sentence.
+Also OLS does \emph{not} give equal weight to all points, such is the nature of \textbf{squared} error.}
 

### 2016-09-20 Paper/figures/figures.tex
JH edits.

--- 2016-09-20 Paper/figures/figures.tex
@@
   \includegraphics[width=.55\columnwidth]{figures/exp4}
-  \caption{A cropped example stimulus from experiment 3.
-We replaced the final 15 points of this dataset with extreme values.
-The overlaid purple trend line represents a robust OLS fit (ignoring the outlier values), while the overlaid green line represents the fit with all points included.
-Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.}
+  \caption{An example stimulus from experiment 3.
+We replaced the final 10 points of this dataset with extreme values.
+The overlaid green trend line represents a robust fit (ignoring the outlier values), while the overlaid orange line represents the standard OLS fit with all points included.
+The purple line represents the average participant response on this stimulus.
+In general, participants' estimates of trend lines were closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.}
   \label{fig:outlier}
@@
 }
+
+\newcommand{\anscombeFig}{
+	\begin{figure*}
+		\centering
+		\includegraphics[width=.75\textwidth]{figures/anscombe}
+		\caption{Anscombe's quartet.
+Each series has nearly identical summary statistics include mean, standard deviation, and linear fit.
+Yet, viewers can disambiguate, through visual inspection, differing patterns and trends.
+We refer to this visual estimation of trend as ``regression by eye.''}
+		\label{fig:anscombe}
+	\end{figure*}
+}
+
+\newcommand{\sigmasFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.95\columnwidth]{figures/sigmas}
+		\caption{The effect of increasing residual bandwidth on stimuli.
+We produced stimuli by creating points on a target trend line (in this case, $f(x) = -0.2x + 0.6$), and then adding residuals drawn evenly from a gaussian with a given bandwidth $\sigma$.
+Larger bandwidths mean more dispersion and so weaker fits.}
+		\label{fig:sigmas}
+	\end{figure}
+}
+
+\newcommand{\expOnesigmasFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.85\columnwidth]{figures/exp1sigmas}
+		\caption{From experiment 1, the effect of increased residual bandwidth on error.
+See Fig.
+\protect\ref{fig:sigmas} for example stimuli at each factor level.
+Absolute error at estimating trend monotonically increases as the goodness of fit decreases.
+Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+}
+		\label{fig:exp1sigmas}
+	\end{figure}
+}
+
+\newcommand{\typesFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.95\columnwidth]{figures/types}
+		\caption{The three types of bivariate visualizations explored in this work: area charts, line charts, and scatterplots.
+The density of points made bar charts and area charts extremely visually similar, and so they were excluded from our experiments.
+Likewise, the error in estimating trends from heatmaps\protect\cite{fuchs2013evaluation} made them unsuitable for the experimental task.}
+		\label{fig:types}
+	\end{figure}
+}
+
+\newcommand{\expTwoTypesFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.85\columnwidth]{figures/exp2types}
+		\caption{From experiment 2, the effect of chart type on signed error.
+See Fig.
+\protect\ref{fig:sigmas} for example stimuli at each factor level.
+Area charts are visually asymmetric, with the area below the line filled in with a color.
+This visual asymmetry results in a form of within-the-bar bias\protect\cite{newman2012bar}, where values in the filled region are perceived as likelier than values outside of it.
+This bias manifests as a consistent underestimation in the intercept of trend lines.
+Other chart types we examined do not have this bias.
+Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+}
+		\label{fig:exp2types}
+	\end{figure}
+}

### 2016-09-20 Paper/proceedings.tex
JH edits.

@@
 
+\usepackage{amsmath}
 % If you want to use todo notes, marginpars etc. during creation of your draft document, you
@@
 These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
-\jeff{These last three sentences seem a bit overlapping / redundant. Plus we really seem to like the word ``systematically''.}
 \end{abstract}
@@
 \section{Introduction}
+\anscombeFig
 
-%motivation here is poor. needs another turn of the crank.
-
-An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}, a set of four bivariate datasets with nearly identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots.
-\jeff{Add figure of Anscombe's quartet...} This example relies on the fact that people have the ability to \emph{perceptually} estimate statistical quantities of interest.
+An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}, a set of four bivariate datasets with nearly identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots (Fig.
+\ref{fig:anscombe}).
+This example relies on the fact that people have the ability to \emph{perceptually} estimate \emph{statistical} quantities of interest.
 Visualization users regularly perform statistical tasks, including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection.
-\jeff{Citations to perceptual work touching on the previous points?
-E.g., scatter plot means, color weaving, others?}
+Recent work has examined the accuracy of these visual inspections for means in scatter plots \cite{gleicher2013perception} and time series data \cite{correll2012comparing}, as well as speculated on the affordances of visualizations in general for supporting this sort of visual estimation of summary statistics \cite{szafir2016four}.
 
@@
 We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers.
-We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and outliers introduce ambiguity in how trends are estimated.
-\jeff{This is itself a bit ambiguous.
-What do you mean by ``introduce ambiguity''?
-Can you be more precise?} These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.
-\jeff{Obligatory design implications remark here?}
+We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and that viewers give low weights to extreme values when estimating trends.
+These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.
+Designers should therefore
 
@@
 As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines).
-As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of key visual proxies.
-\jeff{What do you mean by ``key visual proxies''?
-Be clearer so that readers can follow along.} And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.
-\jeff{Such as? The related work could end on a stronger note if we put forward some concrete hypotheses and note that prior work has not addressed them...}
+As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of visual proxies.
+That is, while viewers are unlikely to be directly estimation correlation \emph{per se}, they are estimating correlation through the perception of visual features such as the envelope of the points, point dispersion, or some other visual feature highly correlated with the statistic of interest.
+
+Prior work in visualization has not directly addressed the capabilities of regression by eye, instead assuming as a given that visual estimates of trend are sufficiently accurate.
+Work in the perception of summary statistics lends credences to this assumption, but we believe that trend estimation can be biased through conscious or unconscious design choices.
+For instance, within-the-bar bias may result in underestimation of trends in bar and area charts, and factors that affect perception of correlation (such as aspect ratio and whitespace) may result in under- or overweighting of outliers.
 
@@
 \subsection{Experimental Interface}
+\typesFig
 
-In each experiment, we presented participants with visualizations of bivariate data and asked them to interactively adjust trend lines to best fit that data.
+We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig.
+\ref{fig:types}).
+In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.
 Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}.
@@
 
+\sigmasFig
+
 For each stimulus we wished to have precise, independent control over relevant statistics such as noise, slope, and sample size, while maintaining the appearance of a ``natural'' distribution of points.
@@
 For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian.
-This set was permuted, and then applied to the points along with a target trend.
+The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig.
+\ref{fig:sigmas}).
+We permuted this set of residuals, and then applied them to points along a target trend.
 As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
@@
 
+\expOnesigmasFig
+
 We designed our first experiment to examine how accurate participants were at estimating the magnitude (slope, amplitude, or curvature) of trends in bivariate visualizations.
-We examined a variety of common bivariate visualizations \jeff{Such as?}.
-In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.
+We examined three types of bivariate visualizations: scatter plots, line graphs, and area chart (with the filled area below the line).In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.
 
 We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend in the points.
-The visualizations were one of a scatter plot, line graph, or area chart (with the filled area below the line).
-The slider parameterized one of three types of trend (linear, quadratic, or trigonometric).
-For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).
-\jeff{To many parentheticals...}
+The slider parameterized one of three types of trend: linear, quadratic, or trigonometric.
+For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line.
+In the case of quadratic trends, this slope was the curvature; for trigonometric fits, the positive/negative amplitude.
 
-Participants saw one of each combination of 3 chart types (scatter plot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, and 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli.
+Participants saw one of each combination of 3 chart types (scatter plot, line graph, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, and 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli.
 We also included an additional 4 validation stimuli otherwise excluded from analysis.
-The type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.
-\jeff{Was this factor random or consistent across subjects?}
+The type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level randomly assigned.
 
@@
 \subsection{Results}
-We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on unsigned error \jeff{absolute error?} in estimation of trend lines.
+We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on absolute error in estimation of trend lines.
 We included participant ids and the actual slope of the trend line as covariates.
@@
 Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}.
-We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, \jeff{p-value?}).
-The trimmed mean \jeff{trimmed how?} of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$.
-Figure \ref{fig:sigma} illustrates these results.
+We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, $p<0.001$).
+The interquartile mean of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$.
+Figure \ref{fig:exp1sigmas} illustrates these results.
 
@@
 Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise-interactions.
-The trimmed mean of absolute estimation error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.
+The interquartile mean of absolute  error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.
 
@@
 The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses ($\Delta=0.01$).
-A Student's T-test failed to support the hypothesis that $mu_{error}\ne0$ ($p=0.70$) \jeff{Test statistic and degrees of freedom?}.
+A Student's T-test failed to support the hypothesis that $mu_{error}\ne0$ ($t(4590) = 0.39$, $p=0.70$).
+
+In general, accuracy at estimation of slope was high, an interquartile mean of $0.06$ for slope values that ranged from $\{-1,1\}$, corresponding to an accuracy of $17\%$ of the actual slope.
 
 \section{Experiment 2: ``Within-the-Bar'' Bias}
+
+\expTwoTypesFig
 
@@
 \begin{enumerate}
-	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatter plots.
+	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line graphs and scatter plots.
 \end{enumerate}
@@
 
-Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatter plots or line graphs}.
+Our results support our first hypothesis: \textbf{participants systematically underestimated the intercept of trends in area charts, but not in scatter plots or line graphs}.
 We observed a significant main effect of graph type on signed error ($F(2,2431)=27$, $p<0.001$).
 A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and line graphs.
-The trimmed mean \jeff{trimmed how?} of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions.
-This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots).
-Figure \ref{fig:outlier} illustrates this result.
+The interquartile mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to an interquartile mean of $0.002$ for the other two conditions.
+This underestimation corresponds to unsigned errors more than twice as large in area charts (interquartile mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots).
+Figure \ref{fig:exp2types} illustrates this result.
 
@@
 Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends.
-However, cognitive biases such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in their estimations \jeff{Need citations in this sentence?}.
-These biases can impact forecasting tasks~\cite{campbell2009anchoring, ji2001culture}.
+However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in prediction tasks ~\cite{campbell2009anchoring, ji2001culture}.
 That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.
@@
 	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
-	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in forecasting tasks suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'') \jeff{cite?}.
+	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in predictions suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'')~\cite{croson2005gambler}.
 We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
@@
 \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}.
-Excluding conditions with zero outliers (and so the two trend lines would be identical), the trimmed mean \jeff{trimmed how?} of unsigned error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$).
-A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($p<0.001$) \jeff{test statistic, dof}.
-However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs.
-$\mu=0.08$, $p<0.001$).
-This indicates that participants are not entirely ignoring outliers, although the small effect size suggestes that participants are giving outliers significantly less weight than OLS regression.
+Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$).
+A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$).
+However, a Welch Two Sample T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs.
+$\mu=0.08$, $t(3808) = 9.8$, $p<0.001$).
+This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression.
 
-\jeff{Whither the other hypotheses?}
+Our results fail to support our second hypothesis.
+\textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}.
+There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$,$F(1,4591)=110$, $p<0.001$).
+However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$,$F(1,4591)=2800$, $p<0.001$).
+Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line increased almost 7-fold, from $0.07$ with 0 outliers, monotonically to $0.54$ with 15 outliers.
+These results indicate that while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.
 
-%There is a confound between the number of outliers and the OLS trend line: since all outliers had approximately the same magnitude, more outliers create more extreme slopes with respect to the robust trend line, and so (if a participant were strictly ignoring outliers), larger differences in unsigned error.
-Similarly, outliers at the ends of the data series will adjust the slope of the OLS line more than outliers in the middle.
-We therefore measured performance as the unsigned error as a percentage of the difference between the slopes of the two types of fit.
-We performed a two-way ANOVA on the effect of outlier number and location on this percentage error measure.
-Participant id was included as a random variable.
+Our results also fail to support our third hypothesis, and seem to support its inverse: \textbf{There was no significant impact of outlier location on performance}.
+There is a confound between outlier location and OLS line: outliers on the ends of the data ranges create larger shifts in slope than outliers in the center, which may only create shifts in intercept.
+To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$).
+A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect ($F(2,3441)=0.63$, $p=0.53$).
 
-%Our results fail to support our second hypothesis.
-\emph{Increasing number of outliers does not increase fidelity to the OLS trend line}.
-A one-way ANCOVA on the effect of outlier numbers greater than 0 on this percentage error showed that outliers were a significant main effect ($F(1,3438)=59$, $p<0.001$).
-However, the percentage error \emph{increases} as the number of outliers increase.
-A post-hoc Tukey's HSD shows that the percentage errors for 10 and 15 outliers are similar (trimmed means are $111\%$ and $112\%$ respectively), they are significantly different from the stimuli with only 5 outliers (trimmed mean of $97\%$).
+%Contrary to our hypothesis, outliers in the middle location resulted in guesses that were closer to the trend line (interquartile mean of interpolation value of $0.66$, compared to $0.24$ for outliers at the beginning location, and $0.32$ for outliers in the final location).
 
@@
 That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.
-\jeff{This point was not so clear in the results discussion of experiment 1. We should discuss (and visualize) the range of accuracies observed there.}
 
@@
 For noisy data, this robustness may be beneficial: people can be relied on to perform filtering operations without explicit guidance.
-However, in other cases, this insensitivity may result in visual estimations at odds with the statistical modeling underpinning other areas of analysis of a particular dataset \jeff{What are you trying to say here?}.
-The emergence of new modes and the shifting of means may not result in a subsequent shift in beliefs about trends in data.
-\jeff{???}
+In other cases, this insensitivity may be undesirable.
+For instance, more complicated automatic analytical processes may use OLS regression or other outlier sensitive techniques as part of their calculations, resulting in a disconnect between human and statistical conclusions at later stages of the sensemaking process.
+Similarly, the human tendency to discount or downweight outliers may result in slow adaptation to new data that does not fit into existing patterns.
 
@@
 Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings (and so associated predispositions or biases about the data) could similarly impact statistical judgments.
-\jeff{Can you clarify this last part?
-And/or provide a concrete example?} Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities.
+For instance, outliers from a source that is perceived as credible could be weighted higher than outliers with more suspect provenance.
+Likewise, preconceptions about the volatility of data from certain domains may color how experts estimate trends.
+Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities.
 Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization~\cite{micallef2012assessing}.
@@
 However, area charts are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations.
-Regression by eye is also less sensitive to outliers than simple statistical regression techniques, resulting in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, can give equal weight all points in a dataset.
-\jeff{Run-on sentence.
-Also OLS does \emph{not} give equal weight to all points, such is the nature of \textbf{squared} error.}
+Regression by eye is also less sensitive to outliers than simple statistical regression techniques.
+This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, which can be heavily influenced by a few extreme values.
 

### 2016-09-20 Paper/figures/figures.tex
Figure dump.

@@
 The purple line represents the average participant response on this stimulus.
-In general, participants' estimates of trend lines were closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.}
+In general, participants' estimates of trend lines were closer to the robust than the non-robust trend; regression by eye tends to downweight outliers compared to OLS.}
   \label{fig:outlier}
+  \end{figure}
+}
+
+\newcommand{\outliersFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.95\columnwidth]{figures/outliers}
+		\caption{The four different outlier numerosities tested in experiment three.
+The overlaid green trend line represents a robust fit (ignoring the outlier values), while the overlaid orange line represents the standard OLS fit with all points included.
+More outliers results in larger divergence between these two types of fit.}
+		\label{fig:outliers}
 	\end{figure}
@@
 		\includegraphics[width=.95\columnwidth]{figures/sigmas}
-		\caption{The effect of increasing residual bandwidth on stimuli.
-We produced stimuli by creating points on a target trend line (in this case, $f(x) = -0.2x + 0.6$), and then adding residuals drawn evenly from a gaussian with a given bandwidth $\sigma$.
+		\caption{The four different Gaussian bandwidths used to generate residuals in this study.
+We produced stimuli by creating points on a target trend line (in this case, $f(x) = -0.2x + 0.6$), and then adding residuals drawn evenly from a Gaussian with a given bandwidth $\sigma$.
 Larger bandwidths mean more dispersion and so weaker fits.}
 		\label{fig:sigmas}
+	\end{figure}
+}
+
+\newcommand{\trendtypesFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.85\columnwidth]{figures/trendtypes}
+		\caption{The three different trend types in this study: linear, quadratic, and trigonometric trends.
+In experiments 1, participants estimated the slope of the linear fits, curvature of the quadratic fits, and amplitude of the trigonometric fits.
+In experiment 2, they estimated the y-intercept of these fits.
+In experiment 3, they estimated only the slope of linear fits.
+Trend lines are displayed here for reference.
+}
+		\label{fig:trendtypes}
 	\end{figure}
@@
 
+\newcommand{\expOnetypesFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.85\columnwidth]{figures/exp1types}
+		\caption{From experiment 1, the effect of different trend types on error.
+See Fig.
+\protect\ref{fig:trendtypes} for example stimuli at each factor level.
+Despite the differing complexity of these types of fit, participant estimates were similar, indicating that regression by eye is capable of non-linear estimates.
+Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+}
+		\label{fig:exp1types}
+	\end{figure}
+}
+
 \newcommand{\typesFig}{
@@
 		\includegraphics[width=.95\columnwidth]{figures/types}
-		\caption{The three types of bivariate visualizations explored in this work: area charts, line charts, and scatterplots.
-The density of points made bar charts and area charts extremely visually similar, and so they were excluded from our experiments.
+		\caption{The three types of bivariate visualizations explored in this work: scatterplots, linegraphs, and area charts.
+The density of points made bar charts and area charts extremely visually similar, and so bar charts were excluded from our experiments.
 Likewise, the error in estimating trends from heatmaps\protect\cite{fuchs2013evaluation} made them unsuitable for the experimental task.}
@@
 }
+
+\newcommand{\expThreeOutliersFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.95\columnwidth]{figures/exp3outliers}
+		\caption{From experiment 3, the effect of numerosity of outliers on absolute error.
+See Fig.
+\protect\ref{fig:outliers} for example stimuli at each factor level.
+As the number of outliers increases, there is increasing divergence between the robust line of best fit (which ignores outliers), and the standard OLS line of best fit (which is sensitive to outliers).
+Estimates become increasingly dissimilar to the OLS fit, indicating that participants downweight of outliers when making estimates of trend.
+However, the increasing error even from the robust trend line indicates that participants still perform some interpolation between the two types of fits.
+Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+}
+		\label{fig:exp3outliers}
+	\end{figure}
+}

### 2016-09-20 Paper/proceedings.tex
Figure dump.

@@
 
+DO SOMETHING
+
 \section{Related Work}
@@
 
+\trendtypesFig
+
 For each stimulus we wished to have precise, independent control over relevant statistics such as noise, slope, and sample size, while maintaining the appearance of a ``natural'' distribution of points.
@@
 We reused these residuals across all different trend types (linear, quadratic, or trigonometric).
+Fig.
+\ref{fig:trendtypes} shows these different trend types on the same set of residuals.
 Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
@@
 \section{Experiment 1: Slope Estimation}
-
-\expOnesigmasFig
 
@@
 \subsection{Results}
+
+\expOnesigmasFig
+
+\expOnetypesFig
+
 We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on absolute error in estimation of trend lines.
 We included participant ids and the actual slope of the trend line as covariates.
+In this experiment, the primary measure of performance was the interquartile mean (or midmean).
+The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions.
+Cleveland \& McGill~\cite{cleveland1984graphical} used the interquartile mean to provide a more robust measure of central tendency for responses from graphical perception studies, where participant error can create long tailed distributions of error.
+This is especially the case for crowdsourced studies, where correcting for data quality issues is perceived as more difficult than in laboratory settings~\cite{heer2010crowdsourcing}.
 
@@
 The interquartile mean of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$.
-Figure \ref{fig:exp1sigmas} illustrates these results.
+Figure \ref{fig:exp1sigmas} illustrates this result.
 
@@
 The interquartile mean of absolute  error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.
+Fig.
+\ref{fig:exp1types} illustrates this result.
 
@@
 \outlierFig
+
+\outliersFig
 OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit.
@@
 We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, outlier-less trend line.
+Fig.
+\ref{fig:outliers} shows examples of this process.
 As with experiment 1, the participants controlled the slope of a rendered trend line with a slider.
@@
 
+\expThreeOutliersFig
+
 Our results support our first hypothesis.
@@
 A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$).
-However, a Welch Two Sample T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs.
+However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs.
 $\mu=0.08$, $t(3808) = 9.8$, $p<0.001$).
 This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression.
+Figure \ref{fig:exp3outliers} illustrates this result.
 

### 2016-09-21 Paper/figures/figures.tex
Dumping more figures, initial stab at implications for design.

@@
 }
+
+\newcommand{\expThreeErrorFig}{
+	\begin{figure}
+		\centering
+		\includegraphics[width=.85\columnwidth]{figures/exp3errorType}
+		\caption{From experiment 3, the absolute error in estimations of trend defined in two ways: either as difference from the robust slope (which excludes outliers), or the OLS slope (which does not).
+Participants were significantly closer to the robust fit, indicating that they were largely insensitive to outliers.
+We include the error from trials with no outliers (and so where the two methods of calculating error agree) for reference.
+Error to the robust line is higher than this standard, indicating that participants were at least performing some interpolation between OLS and robust fits, even if they largely favored the robust fit.
+Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+}
+		\label{fig:exp3errorType}
+	\end{figure}
+}

### 2016-09-21 Paper/proceedings.tex
Dumping more figures, initial stab at implications for design.

@@
 These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.
-Designers should therefore
-
-DO SOMETHING
+Designers should therefore make an informed decision between designing for regression by eye and the explicit annotation of statistical regression information.
+Regression by eye affords quick estimations and flexibility in model selections, but is subject to perceptual biases and inaccuracies in estimation.
+Explicit annotation affords accuracy, but constrains the model space, and adds visual complexity to plots.
 
@@
 \outliersFig
+
 OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit.
@@
 
+\expThreeErrorFig
+
 \expThreeOutliersFig
@@
 $\mu=0.08$, $t(3808) = 9.8$, $p<0.001$).
+Figure \ref{fig:exp3errorType} illustrates this result.
 This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression.
-Figure \ref{fig:exp3outliers} illustrates this result.
+Figure \ref{fig:outlier} shows an example of this weighting.
 
@@
 These results indicate that while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.
+Figure \ref{fig:exp3outliers} illustrates this result.
 

### 2016-09-21 Paper/figures/figures.tex
Jeff's edits and comments.

@@
   \includegraphics[width=.55\columnwidth]{figures/exp4}
-  \caption{An example stimulus from experiment 3.
+  \caption{An example stimulus from Experiment 3.
 We replaced the final 10 points of this dataset with extreme values.
@@
 The purple line represents the average participant response on this stimulus.
-In general, participants' estimates of trend lines were closer to the robust than the non-robust trend; regression by eye tends to downweight outliers compared to OLS.}
+In general, participants' estimates of trend lines were closer to the robust than the non-robust trend; regression by eye tends to down-weight outliers compared to OLS.}
   \label{fig:outlier}
@@
 		\includegraphics[width=.95\columnwidth]{figures/outliers}
-		\caption{The four different outlier numerosities tested in experiment three.
+		\caption{The four different outlier numerosities tested in Experiment 3.
 The overlaid green trend line represents a robust fit (ignoring the outlier values), while the overlaid orange line represents the standard OLS fit with all points included.
@@
   \centering
-  \includegraphics[width=.65\columnwidth]{figures/exp}
+  \includegraphics[width=1.0\columnwidth,trim={0 0 0 10cm},clip]{figures/exp}
   \caption{An example estimation task from our experiments.
@@
 Each series has nearly identical summary statistics include mean, standard deviation, and linear fit.
-Yet, viewers can disambiguate, through visual inspection, differing patterns and trends.
-We refer to this visual estimation of trend as ``regression by eye.''}
+Yet, through visual inspection viewers can disambiguate  differing patterns and trends.
+We refer to this visual estimation of trends as ``regression by eye.''}
 		\label{fig:anscombe}
@@
 		\centering
-		\includegraphics[width=.85\columnwidth]{figures/trendtypes}
+		\includegraphics[width=.95\columnwidth]{figures/trendtypes}
 		\caption{The three different trend types in this study: linear, quadratic, and trigonometric trends.
-In experiments 1, participants estimated the slope of the linear fits, curvature of the quadratic fits, and amplitude of the trigonometric fits.
-In experiment 2, they estimated the y-intercept of these fits.
-In experiment 3, they estimated only the slope of linear fits.
+In Experiment 1, participants estimated the slope of the linear fits, curvature of the quadratic fits, and amplitude of the trigonometric fits.
+In Experiment 2, they estimated the y-intercept of these fits.
+In Experiment 3, they estimated only the slope of linear fits.
 Trend lines are displayed here for reference.
@@
 		\includegraphics[width=.85\columnwidth]{figures/exp1sigmas}
-		\caption{From experiment 1, the effect of increased residual bandwidth on error.
+		\caption{The effect of increased residual bandwidth on error in Experiment 1.
 See Fig.
@@
 Absolute error at estimating trend monotonically increases as the goodness of fit decreases.
-Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+Confidence intervals represent bootstrapped 95\% CIs of interquartile means.
 }
@@
 		\includegraphics[width=.85\columnwidth]{figures/exp1types}
-		\caption{From experiment 1, the effect of different trend types on error.
+		\caption{The effect of different trend types on error in Experiment 1.
 See Fig.
@@
 Despite the differing complexity of these types of fit, participant estimates were similar, indicating that regression by eye is capable of non-linear estimates.
-Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+Confidence intervals represent bootstrapped 95\% CIs of interquartile means.
 }
@@
 		\caption{The three types of bivariate visualizations explored in this work: scatterplots, linegraphs, and area charts.
-The density of points made bar charts and area charts extremely visually similar, and so bar charts were excluded from our experiments.
-Likewise, the error in estimating trends from heatmaps\protect\cite{fuchs2013evaluation} made them unsuitable for the experimental task.}
+The density of points made bar charts and area charts visually similar, and so we excluded bar charts from our experiments.
+Likewise, the error in estimating trends from heatmaps~\protect\cite{fuchs2013evaluation} made them unsuitable for the experimental task.}
 		\label{fig:types}
@@
 		\includegraphics[width=.85\columnwidth]{figures/exp2types}
-		\caption{From experiment 2, the effect of chart type on signed error.
+		\caption{The effect of chart type on signed error in Experiment 2.
 See Fig.
@@
 Area charts are visually asymmetric, with the area below the line filled in with a color.
-This visual asymmetry results in a form of within-the-bar bias\protect\cite{newman2012bar}, where values in the filled region are perceived as likelier than values outside of it.
-This bias manifests as a consistent underestimation in the intercept of trend lines.
+This visual asymmetry results in a form of within-the-bar bias~\protect\cite{newman2012bar}, where values in the filled region are perceived as likelier than values outside of it.
+This bias manifests as a consistent under-estimation in the intercept of trend lines.
 Other chart types we examined do not have this bias.
-Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+Confidence intervals represent bootstrapped 95\% CIs of interquartile means.
 }
@@
 		\includegraphics[width=.95\columnwidth]{figures/exp3outliers}
-		\caption{From experiment 3, the effect of numerosity of outliers on absolute error.
-See Fig.
-\protect\ref{fig:outliers} for example stimuli at each factor level.
+		\caption{The effect of numerosity of outliers on absolute error in Experiment 3.
+See Fig.~\protect\ref{fig:outliers} for example stimuli at each factor level.
 As the number of outliers increases, there is increasing divergence between the robust line of best fit (which ignores outliers), and the standard OLS line of best fit (which is sensitive to outliers).
-Estimates become increasingly dissimilar to the OLS fit, indicating that participants downweight of outliers when making estimates of trend.
+Estimates become increasingly dissimilar to the OLS fit, indicating that participants down-weight outliers when making estimates of trend.
 However, the increasing error even from the robust trend line indicates that participants still perform some interpolation between the two types of fits.
-Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
+Confidence intervals represent bootstrapped 95\% CIs of interquartile means.
 }
@@
 		\includegraphics[width=.85\columnwidth]{figures/exp3errorType}
-		\caption{From experiment 3, the absolute error in estimations of trend defined in two ways: either as difference from the robust slope (which excludes outliers), or the OLS slope (which does not).
+		\caption{The absolute error of trend estimates in Experiment 3, measured as difference from the OLS slope (which includes outliers, row 1) or from the robust slope (which excludes outliers, row 2).
 Participants were significantly closer to the robust fit, indicating that they were largely insensitive to outliers.
-We include the error from trials with no outliers (and so where the two methods of calculating error agree) for reference.
+We also include the absolute error for trials with no outliers for reference.
 Error to the robust line is higher than this standard, indicating that participants were at least performing some interpolation between OLS and robust fits, even if they largely favored the robust fit.
-Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean.
-}
+Confidence intervals represent bootstrapped 95\% CIs of interquartile means.}
 		\label{fig:exp3errorType}

### 2016-09-21 Paper/proceedings.tex
Jeff's edits and comments.

@@
 % so you remain anonymous.
-\def\plaintitle{Regression by Eye: Estimating Trends in Bivariate Visualizations}
+\def\plaintitle{Regression by Eye: Estimating Trends\\in Bivariate Visualizations}
 \def\plainauthor{Removed for Review}
@@
 We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models.
-In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge.
+% In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge.
 These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
+% JEFF: These last two lines largely re-state the preceding sentence. Maybe add a new line to the abstract that summarizes implications for visualization designers or tool makers.
 \end{abstract}
 
-\category{H.5.m.}{Information Interfaces and Presentation
-  (e.g. HCI)}{Miscellaneous}
+\category{H.5.2}{Information Interfaces and Presentation
+  (e.g., HCI)}{Evaluation/Methodology}
 
@@
 
-An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}, a set of four bivariate datasets with nearly identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots (Fig.
+An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}: a set of four bivariate datasets with identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots (Fig.
 \ref{fig:anscombe}).
 This example relies on the fact that people have the ability to \emph{perceptually} estimate \emph{statistical} quantities of interest.
-Visualization users regularly perform statistical tasks, including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection.
-Recent work has examined the accuracy of these visual inspections for means in scatter plots \cite{gleicher2013perception} and time series data \cite{correll2012comparing}, as well as speculated on the affordances of visualizations in general for supporting this sort of visual estimation of summary statistics \cite{szafir2016four}.
+Visualization users regularly perform statistical tasks\,---\,including model selection, identification of outliers, and estimation of summary statistics\,---\,entirely through visual inspection.
+Recent work examines the accuracy of visual estimation of means in scatter plots \cite{gleicher2013perception} and time series data \cite{correll2012comparing}, and speculates on the affordances of visualizations in general for supporting visual estimation of summary statistics \cite{szafir2016four}.
 
@@
 
-Even if designers include all relevant information, the audience may lack the statistical expertise to interpret these values, or may be misled if the data violate modeling assumptions.
+Even if designers include modeling information, the audience may lack the statistical expertise to interpret these values, or may be misled if the data violate modeling assumptions.
 As a further complication, the form of visual encoding may influence viewers' inferences.
@@
 We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers.
-We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and that viewers give low weights to extreme values when estimating trends.
+We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic under-estimation of trend intercept, and that viewers give low weights to extreme values when estimating trends.
 These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.
+
 Designers should therefore make an informed decision between designing for regression by eye and the explicit annotation of statistical regression information.
@@
 Scatter plots are a standard means of visualizing bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}.
-Prior work has confirmed that viewers can make use of scatter plots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}.
-However, the heuristics used when performing these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases.
+Prior work has confirmed that viewers can make use of scatter plots to perform prediction tasks (which tacitly rely on trend estimation) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}.
+However, the heuristics used to perform these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases.
 
 Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}.
-The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}\,---\,wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts.
+The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}, as wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts.
 Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.
@@
 Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatter plots.
-Harrison et al.~\cite{harrison2014ranking} extend this finding to other visualization types, although a re-analysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor.
-Estimation of correlation can also be biased by the choice of scales~\cite{cleveland1982variables}\,---\,the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.
+Harrison et al.~\cite{harrison2014ranking} extend this finding to other visualization types, and a re-analysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor.
+Estimation of correlation can also be biased by the choice of axis scales~\cite{cleveland1982variables}: the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause over-estimation of correlation.
 
@@
 As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of visual proxies.
-That is, while viewers are unlikely to be directly estimation correlation \emph{per se}, they are estimating correlation through the perception of visual features such as the envelope of the points, point dispersion, or some other visual feature highly correlated with the statistic of interest.
+That is, while viewers are unlikely to directly estimate correlation \emph{per se}, they are estimating correlation through the perception of visual features such as the envelope of the points, point dispersion, or some other visual feature(s) highly correlated with the statistic of interest.
 
@@
 Work in the perception of summary statistics lends credences to this assumption, but we believe that trend estimation can be biased through conscious or unconscious design choices.
-For instance, within-the-bar bias may result in underestimation of trends in bar and area charts, and factors that affect perception of correlation (such as aspect ratio and whitespace) may result in under- or overweighting of outliers.
+For instance, within-the-bar bias may result in under-estimation of trends in bar and area charts, and factors that affect perception of correlation (such as aspect ratio and whitespace) may result in up- or down-weighting of outliers.
 
-\section{General Methods}
+\section{General Experimental Methods}
 
@@
 
-In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform.
-We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias for these estimates.
-Crowdsourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}.
-Crowdsourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}.
+In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk.
+We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias.
+Crowdsourced graphical perception experiments have been found to produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}.
 In this section, we describe experimental design aspects shared across all experiments.
@@
 We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
-For validation, certain trials contained the actual (OLS) trend line.
-We excluded the results of participants with poor performance on these validation questions.
-Across the three reported experiments and pilots, we performed 7 such exclusions.
+For validation, certain trials contained the actual (OLS) trend line, which subjects then needed to simply match.
+\jeff{But what if subjects honestly disagreed?
+Did we do this in the presence of outliers, for example?} We excluded the results of participants with poor performance on these validation questions.
+\jeff{What level of error constitutes poor performance?} Across the three reported experiments and pilots, we performed 7 such exclusions.
 We recruited additional participants to replace these excluded subjects.
@@
 We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand.
-We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model space and to reduce the impact of factors such as accuracy of motor movements or heterogeneity of input devices.
+We avoided more expressive designs (such as freeform drawing) in order to constrain the model space and to reduce the impact of factors such as noise from motor movements or heterogeneous input devices.
 
@@
 For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian.
-The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig.
+\jeff{In what way was it discretized?} The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig.
 \ref{fig:sigmas}).
@@
 \ref{fig:trendtypes} shows these different trend types on the same set of residuals.
+
 Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.
@@
 For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term.
-For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes).
+For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes).
 These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.
@@
 We designed our first experiment to examine how accurate participants were at estimating the magnitude (slope, amplitude, or curvature) of trends in bivariate visualizations.
-We examined three types of bivariate visualizations: scatter plots, line graphs, and area chart (with the filled area below the line).In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.
+We examined three types of bivariate visualizations: scatter plots, line graphs, and area charts (with the filled area below the line).
+In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.
 
-We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend in the points.
+We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend.
 The slider parameterized one of three types of trend: linear, quadratic, or trigonometric.
@@
 We anticipated that estimation of these less familiar relationships would therefore be more difficult than the linear case.
-	\item \textbf{Estimations would be unbiased}. That is, there would be no systematic over- or underestimation of trends.
+	\item \textbf{Estimations would be unbiased}. That is, there would be no systematic over- or under-estimation of trends.
 \end{enumerate}
@@
 We included participant ids and the actual slope of the trend line as covariates.
-In this experiment, the primary measure of performance was the interquartile mean (or midmean).
+\jeff{Add a sentence here explicitly defining our error measure.}
+
+In this experiment, the primary measure of performance was the interquartile mean (or midmean) of absolute error.
 The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions.
@@
 Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}.
-Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise-interactions.
-The interquartile mean of absolute  error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.
-Fig.
-\ref{fig:exp1types} illustrates this result.
+Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise interactions.
+The interquartile mean of absolute error of linear fits was $17\%$ of the actual slope, in between a $16\%$ error for quadratic fits, and $18\%$ for trignometric fits (see Fig.
+\ref{fig:exp1types}).
+This result suggests that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.
 
 Our results support our third hypothesis: \textbf{there was no statistically significant bias in estimations}.
-Participants saw an equal number of stimuli with positive and negative trend lines.
+Participants saw a balanced set of positive and negative trends.
 If estimates of these trends were unbiased, we would expect the average signed error to be close to zero.
@@
 
-In general, accuracy at estimation of slope was high, an interquartile mean of $0.06$ for slope values that ranged from $\{-1,1\}$, corresponding to an accuracy of $17\%$ of the actual slope.
+In general, the accuracy of slope estimation was high, with an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an accuracy of $17\%$ of the actual slope.
+\jeff{We use the term ``accuracy'' for two different measurements here, which is confusing.
+Also, a widow line.}
 
@@
 We hypothesized that this bias would likewise occur for regression tasks when using visually asymmetric visualizations such as area charts.
-The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} underestimation in trends; decreasing the slope would cause underestimation at the beginning of the plot but not the end, and vice versa.
+The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa.
 We therefore designed this experiment to elicit estimates of the y-intercepts of trends.
-A within-the-bar bias would then appear as systematic underestimation of intercept in area charts.
+A within-the-bar bias would then appear as systematic under-estimation of intercept in area charts.
 
@@
 A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and line graphs.
-The interquartile mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to an interquartile mean of $0.002$ for the other two conditions.
-This underestimation corresponds to unsigned errors more than twice as large in area charts (interquartile mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots).
+The interquartile mean of the signed error of estimations made with area charts was an under-estimation of $-0.02$, compared to an interquartile mean of $0.002$ for the other two conditions.
+This under-estimation corresponds to unsigned errors more than twice as large in area charts (interquartile mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots).
 Figure \ref{fig:exp2types} illustrates this result.
 
-\section{Experiment 3. Estimation involving Outliers}
+\section{Experiment 3: Estimation Involving Outliers}
 
@@
 Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods.
-Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}.
+Visual inspection can accurately identify certain classes of outliers~\cite{albers2014task}.
 Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends.
-However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in prediction tasks ~\cite{campbell2009anchoring, ji2001culture}.
-That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.
+However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values will continue) can impact how viewers categorize and utilize outliers in prediction tasks~\cite{campbell2009anchoring, ji2001culture}.
+In other words, regression by eye may not uniformly weight outliers, depending on their position in the plot.
 
-This experiment was largely identical to experiment 1 above, except we designated 0, 5, 10, or 15 points at the very beginning, first third, or end of the series as outliers, and placed them randomly in the top or bottom 10\% of the visual area of the visualization (whichever was farthest from the trend line).
+This experiment was largely identical to Experiment 1, except we designated 0, 5, 10, or 15 points at the very beginning, first third, or end of the series as outliers.
+We vertically positioned outliers within the top or bottom 10\% of the visual area of the visualization (whichever was farthest from the trend line), and added random jitter.
 We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, outlier-less trend line.
-Fig.
-\ref{fig:outliers} shows examples of this process.
-As with experiment 1, the participants controlled the slope of a rendered trend line with a slider.
+Figure \ref{fig:outliers} shows examples of this process.
+
+As in Experiment 1, the participants controlled the slope of a rendered trend line with a slider.
 However, rather than being placed such that $f(0.5)=0.5$, we offset the trend line to the intersection of the robust (without outliers) and non-robust (including outliers) trend line.
 This allowed participants to express both types of fit with the same slider interaction, while affording estimations beyond a simple interpolation of both trends.
-This expressiveness is only afforded by linear fits of, and so we did not consider quadratic or trigonometric trends in this experiment.
+We limited the stimuli to linear fits only, excluding quadratic and trigonometric trends from this experiment.
 
 Similar to the previous experiments, participants saw one of each of the three graph types, slope, and Gaussian residual bandwidth.
-The four outlier quantities $\{0,5,10,15\}$ was an additional factor.
+\jeff{Awkward grammar for previous sentence.} The four outlier quantities $\{0,5,10,15\}$ was an additional factor.
 To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively.
@@
 \begin{enumerate}
-	\item \textbf{Participant estimation would be closer to a robust trend line ignoring outliers, than the non-robust OLS trend line.} That is, we assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
+	\item \textbf{Participant estimation would hew closer to a robust trend line ignoring outliers, rather than the OLS trend line.} We assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
 	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
-	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in predictions suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'')~\cite{croson2005gambler}.
+	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in predictions suggests that viewers may give more credence to more recent outliers (as indicative as new anchor points, or of an emerging ``streak'')~\cite{croson2005gambler}.
 We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
@@
 \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}.
-There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$,$F(1,4591)=110$, $p<0.001$).
-However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$,$F(1,4591)=2800$, $p<0.001$).
+There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$, $F(1,4591)=110$, $p<0.001$).
+However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$, $F(1,4591)=2800$, $p<0.001$).
 Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line increased almost 7-fold, from $0.07$ with 0 outliers, monotonically to $0.54$ with 15 outliers.
@@
 In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied.
-Yet, our results point to regression by eye as a robust and accurate method for estimating trends in bivariate data.
+Yet, our results point to regression by eye as a robust and reasonably accurate method for estimating trends in bivariate data.
 Participants from a variety of backgrounds and levels of self-reported graphical and statistical expertise were capable of estimating both the slope and intercept of both linear and non-linear trends.
@@
 Area charts are visually asymmetrical: the area below the line is filled in, and the area above it is not.
-We found that this asymmetry creates a within the bar bias: a systematic underestimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area.
+We found that this asymmetry creates a within the bar bias: a systematic under-estimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area.
 Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.
@@
 
-Finally, we focused on a simple form of regression, OLS, as our standard for measuring accuracy.
+Finally, we focused on a simple form of regression, ordinary least squares (OLS), as our standard for measuring accuracy.
 While our data were constructed to satisfy the assumptions of OLS (with the intentional exception of our outlier experiment), in most real world scenarios OLS is just one tool of many, and analysts must exercise judgment when determining how to fit their data.
@@
 First, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye.
-Can sufficient information promote caution in judgments of trend, or do initial visual estimations have too much weight to be overcome?
+Can sufficient information promote caution in judgments of trend?
 Second, we wish to examine the impact of different rhetorical framings and presentations on regression by eye.
-Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings (and so associated predispositions or biases about the data) could similarly impact statistical judgments.
+Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings could similarly impact statistical judgments.
 For instance, outliers from a source that is perceived as credible could be weighted higher than outliers with more suspect provenance.
@@
 
-In this paper, we examine the ability of viewers of visualizations to perform \emph{regression by eye}\,---\,the visual estimation of trends in bivariate visualizations.
-We show that even viewers without statistical training can reliably estimate both linear and non-linear trends in visualizations such as scatter plots and line graphs.
-However, area charts are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations.
-Regression by eye is also less sensitive to outliers than simple statistical regression techniques.
-This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, which can be heavily influenced by a few extreme values.
+In this paper, we examine the ability of visualization viewers to perform \emph{regression by eye}: the visual estimation of trends in bivariate visualizations.
+We show that viewers without statistical training can reliably estimate both linear and non-linear trends in charts such as scatter plots and line graphs.
+However, area graphs are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations.
+Regression by eye is also less sensitive to outliers than standard least-squares regression.
+This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by regression methods that can be heavily influenced by a few extreme values.
 

### 2016-09-21 Paper/proceedings.tex
Addressed Jeff Edits

@@
 We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
-For validation, certain trials contained the actual (OLS) trend line, which subjects then needed to simply match.
-\jeff{But what if subjects honestly disagreed?
-Did we do this in the presence of outliers, for example?} We excluded the results of participants with poor performance on these validation questions.
-\jeff{What level of error constitutes poor performance?} Across the three reported experiments and pilots, we performed 7 such exclusions.
+For validation, additional trials, without ambiguities such as outliers or design differences, contained the actual (OLS) trend line, which subjects then needed to simply match.
+We excluded the results of participants with average error greater than 0.2 (20 times the step size of the slider) on these validation questions.
+Across the three reported experiments and pilots, we performed 7 such exclusions.
 We recruited additional participants to replace these excluded subjects.
@@
 
-For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian.
-\jeff{In what way was it discretized?} The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig.
+For each point set, we created a set of residual values, sampled evenly from a Gaussian.
+The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig.
 \ref{fig:sigmas}).
@@
 
-We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on absolute error in estimation of trend lines.
+We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on error in estimation of trend lines.
 We included participant ids and the actual slope of the trend line as covariates.
-\jeff{Add a sentence here explicitly defining our error measure.}
+We defined error as the absolute difference between the slope of the OLS trend line, and the slope of the participant estimate.
 
@@
 
-In general, the accuracy of slope estimation was high, with an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an accuracy of $17\%$ of the actual slope.
-\jeff{We use the term ``accuracy'' for two different measurements here, which is confusing.
-Also, a widow line.}
+In general, the absolute error of slope estimation was low, with an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slope.
 
@@
 
-Similar to the previous experiments, participants saw one of each of the three graph types, slope, and Gaussian residual bandwidth.
-\jeff{Awkward grammar for previous sentence.} The four outlier quantities $\{0,5,10,15\}$ was an additional factor.
+Similar to the previous experiments, participants saw one of each combination of factors: the three graph types, eight slopes, and four Gaussian residual bandwidths.
+The four outlier quantities $\{0,5,10,15\}$ was an additional factor.
 To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively.

### 2016-09-21 Paper/proceedings.tex
Attempt at a "see, people are pretty good at this" paragraph.

@@
 
+%MC Packages
 \usepackage{amsmath}
+\usepackage{cite}
+
 % If you want to use todo notes, marginpars etc. during creation of your draft document, you
@@
 Harrison et al.~\cite{harrison2014ranking} extend this finding to other visualization types, and a re-analysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor.
-Estimation of correlation can also be biased by the choice of axis scales~\cite{cleveland1982variables}: the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause over-estimation of correlation.
+Estimation of correlation can also be biased by the choice of axis scales~\cite{cleveland1982variables}: the whitespace changes introduced by expanding the scale of the axes can cause over-estimation of correlation.
 
@@
 
+For each point set, we placed 100 points along a particular trend, regularly spaced in x.
+This both affords the use of linegraphs and area charts for displaying this data (as they require that points be one to one), and aligns well with time series data, a common domain for estimating (temporal) trends and relationships.
 For each point set, we created a set of residual values, sampled evenly from a Gaussian.
@@
 \ref{fig:sigmas}).
-We permuted this set of residuals, and then applied them to points along a target trend.
+We permuted this set of residuals, and then applied them to the original points.
 As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
@@
 We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, $p<0.001$).
-The interquartile mean of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$.
+The interquartile mean of absolute error increased monotonically with this bandwidth, from $0.02$ of the actual slope of the trend line when the bandwidth was $0.05$, to $0.12$ when the bandwidth was $0.20$.
 Figure \ref{fig:exp1sigmas} illustrates this result.
@@
 Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise interactions.
-The interquartile mean of absolute error of linear fits was $17\%$ of the actual slope, in between a $16\%$ error for quadratic fits, and $18\%$ for trignometric fits (see Fig.
+The interquartile mean of absolute error of was $0.06$ for linear and quadratic fits, and $0.07$ for trigonometric fits (see Fig.
 \ref{fig:exp1types}).
@@
 
-In general, the absolute error of slope estimation was low, with an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slope.
+%The average absolute error of slope estimation across all conditions had an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slopes.
+As a point of comparison, the interquartile mean of the absolute log error, calculated as per Cleveland \& McGill~\cite{cleveland1984graphical} ($log_2 |\text{error}| + \frac{1}{8}$) was $2.4$ across all conditions.
+This is comparable to the error rates of many other graphical perception tasks as observed in both lab and crowdsourced studies.
+For instance, Heer \& Bostock~\cite{heer2010crowdsourcing}  report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$).
+This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with the expected accuracy in comparing individual values in visualizations.
 
@@
 We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, outlier-less trend line.
-Figure \ref{fig:outliers} shows examples of this process.
+We refer to the original trend line, estimated without the presence of outliers, as the \textbf{robust trend line}.
+We refer to the re-estimated trend, which takes into account the added outliers, as the \textbf{OLS trend line}.
+Figure \ref{fig:outliers} shows examples of this process, and the corresponding changes in trend line.
 
@@
 
-Similar to the previous experiments, participants saw one of each combination of factors: the three graph types, eight slopes, and four Gaussian residual bandwidths.
-The four outlier quantities $\{0,5,10,15\}$ was an additional factor.
-To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively.
+Participants saw one of each combination of factors: the three graph types, eight slopes, and four Gaussian residual bandwidths.
+The four outlier quantities $\{0,5,10,15\}$ were an additional factor.
+To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level randomly apportioned to half and one third of the stimuli respectively.
 This resulted in 96 total stimuli, with an additional 4 validation stimuli otherwise excluded from analysis, in line with prior experiments.
@@
 Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are fixed.
+It is also likely that there are many visual features, such as the convex hull of points, or the numerosity of point clouds, that are acting as visual proxies for estimation of the trend.
+Given the unimodal error function and uniform density of our experimental stimuli, these proxies are useful for estimating the trend line.
+However, in datasets where internal densities of points may be skewed, or the envelope of point clouds non-informative for estimating central tendency, these visual proxies may introduce biases and confounds in visual estimations.
 

### 2016-09-21 Paper/figures/figures.tex
Typo pass.

@@
 Each series has nearly identical summary statistics include mean, standard deviation, and linear fit.
-Yet, through visual inspection viewers can disambiguate  differing patterns and trends.
+Yet, through visual inspection, viewers can disambiguate  differing patterns and trends.
 We refer to this visual estimation of trends as ``regression by eye.''}
@@
 		\includegraphics[width=.95\columnwidth]{figures/types}
-		\caption{The three types of bivariate visualizations explored in this work: scatterplots, linegraphs, and area charts.
+		\caption{The three types of bivariate visualizations explored in this work: scatter plots, line graphs, and area charts.
 The density of points made bar charts and area charts visually similar, and so we excluded bar charts from our experiments.

### 2016-09-21 Paper/proceedings.tex
Typo pass.

@@
 Knowing the strengths and limitations of such ``estimation by eye'' is therefore important for designers of data visualizations seeking to communicate statistical quantities, especially to a general audience.
-On the one hand, visual estimations may be \emph{inaccurate} for the use cases intended by a designer, or they may be \emph{biased}, leading to systematic over- or under-estimations.
+On the one hand, visual estimations may be too \emph{inaccurate} for the use cases intended by a designer, or they may be \emph{biased}, leading to systematic over- or under-estimations.
 On the other hand, while visual estimation lacks the precision of formal statistics, it may be relatively unencumbered by modeling assumptions.
@@
 However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}.
-A further difficulty is that tasks requiring estimation of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can result in a multitude of potential valid responses).
+A further difficulty is that tasks requiring estimation of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can result in a multitude of potentially valid responses).
 
@@
 
-We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $\mu_{age}= 33.2$, $\sigma_{age}=8.8$).
+We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $M_{age}= 33.2$, $SD_{age}=8.8$).
 Across the subject pool, 10 reported having graduate or professional degrees, 69 college degrees, 40 at least some college, and 25 high school diplomas.
@@
 For each point set, we placed 100 points along a particular trend, regularly spaced in x.
-This both affords the use of linegraphs and area charts for displaying this data (as they require that points be one to one), and aligns well with time series data, a common domain for estimating (temporal) trends and relationships.
+This both affords the use of line graphs and area charts for displaying this data (as they require that points be one to one), and aligns well with time series data, a common domain for estimating (temporal) trends and relationships.
 For each point set, we created a set of residual values, sampled evenly from a Gaussian.
@@
 We permuted this set of residuals, and then applied them to the original points.
-As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend.
+As heteroskedasticity introduced through permutation could alter the trend away from the target trend, we performed rejection sampling to ensure that the slope of the trend of the resulting points was within $10^{-7}$ of the target.
 We reused these residuals across all different trend types (linear, quadratic, or trigonometric).
@@
 
-Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}.
+Our results support our first hypothesis: \textbf{larger residuals result in less accuracy at regression by eye}.
 We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, $p<0.001$).
@@
 %The average absolute error of slope estimation across all conditions had an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slopes.
-As a point of comparison, the interquartile mean of the absolute log error, calculated as per Cleveland \& McGill~\cite{cleveland1984graphical} ($log_2 |\text{error}| + \frac{1}{8}$) was $2.4$ across all conditions.
-This is comparable to the error rates of many other graphical perception tasks as observed in both lab and crowdsourced studies.
+Prior work in graphical perception often measures the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as ($log_2 |\text{error}| + \frac{1}{8}$).
+As a point of comparison the interquartile mean of the absolute log error, calculated as per  was $2.4$ across all conditions in this experiment.
+Differences in methodology and measurement discourage statistical inferences comparing this value with those in similar experiments.
+Nevertheless, error rates for regression by eye are at least in the same magnitude as those observed in other graphical perception tasks in both lab and crowdsourced studies.
 For instance, Heer \& Bostock~\cite{heer2010crowdsourcing} report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$).
@@
 
-We hypothesized that this bias would likewise occur for regression tasks when using visually asymmetric visualizations such as area charts.
+We hypothesized that this bias would likewise occur in regression by eye when using visually asymmetric visualizations such as area charts.
 The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa.
@@
 \begin{enumerate}
-	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line graphs and scatter plots.
+	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends in area charts, as opposed to line graphs and scatter plots.
 \end{enumerate}
@@
 \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}.
-Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$).
+Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to when calculated from the robust trend line ($0.08$).
 A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$).
@@
 This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression.
-Figure \ref{fig:outlier} shows an example of this weighting.
+Figure \ref{fig:outlier} shows an comparison of robust, participant, and OLS trend lines.
 
@@
 However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$, $F(1,4591)=2800$, $p<0.001$).
-Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line increased almost 7-fold, from $0.07$ with 0 outliers, monotonically to $0.54$ with 15 outliers.
-These results indicate that while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.
+Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line monotonically increased almost 7-fold, from $0.07$ with 0 outliers to $0.54$ with 15 outliers.
+These results indicate that, while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.
 Figure \ref{fig:exp3outliers} illustrates this result.
 
-Our results also fail to support our third hypothesis, and seem to support its inverse: \textbf{There was no significant impact of outlier location on performance}.
+Our results also fail to support our third hypothesis: \textbf{There was no significant impact of outlier location on performance}.
 There is a confound between outlier location and OLS line: outliers on the ends of the data ranges create larger shifts in slope than outliers in the center, which may only create shifts in intercept.
 To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$).
-A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect ($F(2,3441)=0.63$, $p=0.53$).
+A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect of outlier location on interpolative value ($F(2,3441)=0.63$, $p=0.53$).
 
@@
 
-Likewise, viewers do not give the same weight to outliers as OLS regression, but they do not ignore them either.
+Likewise, while viewers do not give the same weight to outliers as OLS regression, they do not ignore them either.
 For noisy data, this robustness may be beneficial: people can be relied on to perform filtering operations without explicit guidance.
@@
 In actual regression by eye, the viewer may simultaneously engage in multiple types of estimation: choosing a particular type of fit, ignoring outliers, and estimating the parameters of the chosen model.
-Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are fixed.
+Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are fixed \emph{a priori}.
 It is also likely that there are many visual features, such as the convex hull of points, or the numerosity of point clouds, that are acting as visual proxies for estimation of the trend.
@@
 Regression by eye is also less sensitive to outliers than standard least-squares regression.
-This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by regression methods that can be heavily influenced by a few extreme values.
+This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by regression methods, that can be heavily influenced by a few extreme values.
 

### 2016-09-21 Paper/proceedings.tex
Submitted version.

@@
 %The average absolute error of slope estimation across all conditions had an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slopes.
-Prior work in graphical perception often measures the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as ($log_2 |\text{error}| + \frac{1}{8}$).
-As a point of comparison the interquartile mean of the absolute log error, calculated as per  was $2.4$ across all conditions in this experiment.
+Prior work in graphical perception often measures performance as the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as $\log_2 (|\text{error}| + \frac{1}{8}$).
+As a point of comparison, the interquartile mean of the absolute log error across all conditions in this experiment was $2.4$.
 Differences in methodology and measurement discourage statistical inferences comparing this value with those in similar experiments.

### 2016-11-20 Paper/Rebuttal.txt
Added draft rebuttal

We thank the reviewers for their thoughtful feedback. We are excited to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission with greater utility to the research audience.

If accepted, in addition to the more minor fixes for issues presented by the reviewers, we intend to revise three key areas, as per the guidance of the meta review:

1) Clarification of the statistics used in the paper:

We stand by our decision to use interquartile means and bootstrapped confidence intervals. In addition to [8], these method for showing effect sizes have also been used in more recent graphical perception papers, such as [29]. We will expand our justification for these techniques in the paper, and  clarify our methodology.

IQMs: 
[8] (in section 4.4) uses “the means of the midmeans [IQM] of the log absolute errors” in their charts, including those in the caterpillar plot in [8] Fig. 16 (which R3 is correct in citing as a model for the figures in this paper). R3 correctly notes that the IQM has some disadvantages. In particular, the distribution of such values tends to violate the assumptions of normality of sampling error, which is key for tests such as t-tests and ANOVAs. In the paper, we use the IQM only for plots and for reporting effect sizes; for ANOVAs and other inferential tests, we test the standard mean. We will clarify this distinction in the paper.

Bootstrapped CIs:
Bootstrapped c.i.s appear frequently in graphical perception work, e.g. in [8,20,29]. [8] (also in section 4.4) justifies the use of bootstrapping as a robust and non-parametric way of creating c.i.s.

While we reiterate that the p-values reported in the paper were generated from standard means rather than the IQM, a more troubling issue with IQMs is that they reduce the variance of the variables of interest, and so can artificially inflate effect sizes, or create effects where none exist with other measures of central tendency. To set both meta, R3, and our own minds at ease, we generated CIs using a more standard approach, a two-tailed 95% t-confidence interval about the mean. While these CIs result in smaller effect sizes (due to higher variance), they do not alter any of the main results of the paper.

e.g., Figure 6 would be:
sigma, mean(absolute error) +/- 95%t-ci
0.05,  0.04 +/- (0.004)
0.10,  0.10 +/- (0.005)
0.15,  0.15 +/- (0.006)
0.20,  0.23 +/- (0.01)

Figure 8 would be:
graph type, mean(error) +/- 95%t-ci
area,  -0.01 +/- (0.006)
line,  0.01 +/- (0.005)
scatterplot, 0.005 +/- (0.003)

Figure 11 would be:
error type, mean(absolute error) +/- 95%t-ci
w/r/t OLS fit, 0.4 +/- 0.009
w/r/t robust fit, 0.15 +/- 0.007
no outliers, 0.1 +/- 0.006

and so on. While the data are noisier using this procedure, the trends (monotonic increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts but not other chart types, participants are closer to the robust fit than the OLS fit, etc.) remain.

2) Additional detail about experimental methods:

We will include or clarify study details (number of exclusions, re-structuring of the methods section to make the participant information clearer, etc.) asked for by the reviewers.

Additionally, while we have already included our study data in our supplemental materials, on acceptance, we would make public our (currently private) repository containing not only data tables and analyses, but also experimental stimuli and instruments. In the main body of the paper, we will include figures with more example stimuli. We can include additional figures (such as violin plots of per-subject IQMs, or other non-parametric visualizations of effect size) in the online supplement as well.

R4 and R5 mention, as an alternative study design, un-constrained input of trend lines, rather than our study design (which is more concerned with fitting a given model, linear or otherwise, than the freeform estimation of a fit, which may or may not correspond to any of the categories of provided models). For the targeted research questions in our study, we believe our more parsimonious study design is justified. However, our group is actively researching the design and implementation of experimentally tractable freeform solicitation techniques (for instance, the “draw your guess” NYT example mentioned by R5). We will clarify the distinction between these two types of visual regression, and cite work in this area that has been published since our initial submission.

3) Better situation of this work within the CHI community, and better communication of its utility:

While many graphical perception works have had a good reception at CHI (Fuchs et al “Evaluation of Alternative Glyph Designs for Time Series Data in a Small Multiple Setting” and [10] were presented at CHI 2013 and 2012, respectively), we believe that we can improve the perceived fit of this work at CHI through the inclusion of an explicit implications for design subsection, as part of the conclusion (as per R5). We believe the strength of our results offers guidance to designers in information visualization: in particular, it seems to suggest that area charts (which are a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends. Similarly, it suggest that standard trend lines (which are OLS lines by default in programs like Excel and standard statistical toolkits) may not appear to be credible if there are many outliers in the data; in such cases, it appears that viewers can be trusted to perform the robust fits visually.

This issue of trust in the viewer (or the stats!) to be “well behaved” when it comes to estimating statistical properties, or making use of visualizations, is central to how we developed our research questions for this paper, and we will expand upon it in our introduction (particular our Anscombe quartet example, as per R5), and in scoping our claimed contributions.

We hope that this rebuttal addresses the main concerns brought up by the reviewers, and would like to thank them again for their helpful and thorough comments. As the revisions we propose above are largely to the argumentation and presentation of the paper, we are confident that we can carry them out within the CHI revision cycle, and apply them to the final paper.

### 2016-11-20 Paper/Rebuttal.txt
Jeff's revision.

@@
 We thank the reviewers for their thoughtful feedback.
-We are excited to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission with greater utility to the research audience.
+We hope to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission with greater utility.
+If accepted (in addition to other, more minor changes suggested by the reviewers), as per the meta-review we will revise three key areas:
 
-If accepted, in addition to the more minor fixes for issues presented by the reviewers, we intend to revise three key areas, as per the guidance of the meta review:
+1) Clarification of the statistical methods:
 
-1) Clarification of the statistics used in the paper:
-
-We stand by our decision to use interquartile means and bootstrapped confidence intervals.
-In addition to [8], these method for showing effect sizes have also been used in more recent graphical perception papers, such as [29].
+We stand by our decision to present midmeans and bootstrapped confidence intervals.
+In addition to [8], these methods for showing effect sizes have also been used in more recent graphical perception papers [17, 29].
 We will expand our justification for these techniques in the paper, and clarify our methodology.
 
-IQMs: 
-[8] (in section 4.4) uses “the means of the midmeans [IQM] of the log absolute errors” in their charts, including those in the caterpillar plot in [8] Fig.
-16 (which R3 is correct in citing as a model for the figures in this paper).
-R3 correctly notes that the IQM has some disadvantages.
-In particular, the distribution of such values tends to violate the assumptions of normality of sampling error, which is key for tests such as t-tests and ANOVAs.
-In the paper, we use the IQM only for plots and for reporting effect sizes; for ANOVAs and other inferential tests, we test the standard mean.
+Bootstrapped CIs:
+Bootstrapped CIs appear frequently in graphical perception work [8, 17, 29]. Prior work ([8], section 4.4) justifies the use of bootstrapping as a robust and non-parametric way of creating CIs.
+
+Midmeans (aka Interquartile Means, or IQMs):
+Prior work [8, 17] indeed uses "the means of the midmeans of the log absolute errors" in charts, including Fig.
+16 in [8] (which R3 correctly cites as a model for figures in our paper).
+R3 notes some disadvantages of IQMs: the distribution of such values tends to violate the assumptions of normality of sampling error, which is key for t-tests and ANOVAs.
+In the paper, we use IQMs only for plots and for reporting effect sizes; for ANOVAs and other inferential tests, we use standard means.
 We will clarify this distinction in the paper.
 
-Bootstrapped CIs:
-Bootstrapped c.i.s appear frequently in graphical perception work, e.g. in [8,20,29]. [8] (also in section 4.4) justifies the use of bootstrapping as a robust and non-parametric way of creating c.i.s.
-
-While we reiterate that the p-values reported in the paper were generated from standard means rather than the IQM, a more troubling issue with IQMs is that they reduce the variance of the variables of interest, and so can artificially inflate effect sizes, or create effects where none exist with other measures of central tendency.
-To set both meta, R3, and our own minds at ease, we generated CIs using a more standard approach, a two-tailed 95% t-confidence interval about the mean.
+While we reiterate that the p-values reported in the paper come from standard means (not IQMs), it is true that IQMs reduce the variance of the variables of interest, and so may artificially inflate effect sizes or create effects unseen with other measures of central tendency.
+In response to this concern, we generated CIs using a more standard approach: a two-tailed 95% t-confidence interval of the standard mean.
 While these CIs result in smaller effect sizes (due to higher variance), they do not alter any of the main results of the paper.
 
-e.g., Figure 6 would be:
+Figure 6 becomes:
 sigma, mean(absolute error) +/- 95%t-ci
@@
 
-Figure 8 would be:
+Figure 8 becomes:
 graph type, mean(error) +/- 95%t-ci
@@
 
-Figure 11 would be:
+Figure 11 becomes:
 error type, mean(absolute error) +/- 95%t-ci
@@
 
-and so on.
-While the data are noisier using this procedure, the trends (monotonic increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts but not other chart types, participants are closer to the robust fit than the OLS fit, etc.) remain.
+...and so on.
+While estimates are noisier under this procedure, the results remain unchanged: monotonic increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts but not other chart types, participants are closer to the robust fit than the OLS fit, etc.
 
@@
 
-We will include or clarify study details (number of exclusions, re-structuring of the methods section to make the participant information clearer, etc.) asked for by the reviewers.
+We will include or clarify study details asked for by reviewers, including the number of exclusions, and re-structuring of the methods section to make the participant information clearer.
 
-Additionally, while we have already included our study data in our supplemental materials, on acceptance, we would make public our (currently private) repository containing not only data tables and analyses, but also experimental stimuli and instruments.
-In the main body of the paper, we will include figures with more example stimuli.
-We can include additional figures (such as violin plots of per-subject IQMs, or other non-parametric visualizations of effect size) in the online supplement as well.
+Additionally, though we did submit our study data as supplemental material, on acceptance, we would make public our full (currently private) repository containing not only data tables and analyses, but also experimental stimuli and instruments.
+In the main body of the paper, we will add figures with more example stimuli.
+We can also include additional figures (such as violin plots of per-subject IQMs, or other non-parametric visualizations of effect size) in the online supplement.
 
-R4 and R5 mention, as an alternative study design, un-constrained input of trend lines, rather than our study design (which is more concerned with fitting a given model, linear or otherwise, than the freeform estimation of a fit, which may or may not correspond to any of the categories of provided models).
+As an alternative to our study design, both R4 and R5 mention un-constrained input of trend lines (which we had also initially considered).
+Our design is more concerned with fitting a given model (linear or otherwise) than a freeform "sketch" of a fit, which may or may not correspond to any of the categories of provided models -- and may also suffer from additional noise such as drawing "jitter" or "satisficing" by subjects due to the higher cost of revising drawings.
 For the targeted research questions in our study, we believe our more parsimonious study design is justified.
-However, our group is actively researching the design and implementation of experimentally tractable freeform solicitation techniques (for instance, the “draw your guess” NYT example mentioned by R5).
-We will clarify the distinction between these two types of visual regression, and cite work in this area that has been published since our initial submission.
+However, we know of others actively researching the design and implementation of experimentally tractable freeform solicitation techniques (for instance, the “draw your guess” NYT example mentioned by R5).
+We will clarify the distinction between these two types of visual regression, citing work in this area that has been published since our initial submission.
 
-3) Better situation of this work within the CHI community, and better communication of its utility:
+3) Better situate the work in the CHI community, and better communicate its utility:
 
-While many graphical perception works have had a good reception at CHI (Fuchs et al “Evaluation of Alternative Glyph Designs for Time Series Data in a Small Multiple Setting” and [10] were presented at CHI 2013 and 2012, respectively), we believe that we can improve the perceived fit of this work at CHI through the inclusion of an explicit implications for design subsection, as part of the conclusion (as per R5).
-We believe the strength of our results offers guidance to designers in information visualization: in particular, it seems to suggest that area charts (which are a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends.
-Similarly, it suggest that standard trend lines (which are OLS lines by default in programs like Excel and standard statistical toolkits) may not appear to be credible if there are many outliers in the data; in such cases, it appears that viewers can be trusted to perform the robust fits visually.
+While many graphical perception works have had a welcome reception at CHI (e.g., [10, 17]), we believe that we can improve the perceived fit of this work at CHI (as per R5) through the inclusion of an explicit implications for design subsection of the conclusion.
+We believe our results offer guidance to designers in information visualization.
+In particular, they suggest that area charts (which are a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends.
+Similarly, they suggest that standard trend lines (which are OLS lines by default in programs like Excel and other common analysis tools) may not appear to be credible if there are many outliers in the data; in such cases, it appears that viewers can be trusted to perform the robust fits visually.
 
-This issue of trust in the viewer (or the stats!) to be “well behaved” when it comes to estimating statistical properties, or making use of visualizations, is central to how we developed our research questions for this paper, and we will expand upon it in our introduction (particular our Anscombe quartet example, as per R5), and in scoping our claimed contributions.
+This issue of trust in the viewer (or the stats!) to be "well behaved" when it comes to estimating statistical properties, or making use of visualizations, is central to how we developed our research questions for this paper, and we will expand upon it in our introduction (particular our Anscombe quartet example, as per R5), and in scoping our claimed contributions.
 
-We hope that this rebuttal addresses the main concerns brought up by the reviewers, and would like to thank them again for their helpful and thorough comments.
-As the revisions we propose above are largely to the argumentation and presentation of the paper, we are confident that we can carry them out within the CHI revision cycle, and apply them to the final paper.
+We hope our rebuttal addresses the main reviewer concerns, and would like to thank the reviewers again for their helpful and thorough comments.
+As the revisions we propose above largely concern argumentation and presentation issues, we are confident that we can carry them out within the CHI revision cycle.

### 2016-11-20 Paper/Rebuttal.txt
Trimming rebuttal down to 5000 character limit

@@
 We thank the reviewers for their thoughtful feedback.
-We hope to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission with greater utility.
+We hope to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission.
 If accepted (in addition to other, more minor changes suggested by the reviewers), as per the meta-review we will revise three key areas:
@@
 
-While we reiterate that the p-values reported in the paper come from standard means (not IQMs), it is true that IQMs reduce the variance of the variables of interest, and so may artificially inflate effect sizes or create effects unseen with other measures of central tendency.
-In response to this concern, we generated CIs using a more standard approach: a two-tailed 95% t-confidence interval of the standard mean.
-While these CIs result in smaller effect sizes (due to higher variance), they do not alter any of the main results of the paper.
+While we reiterate that the p-values reported in the paper come from standard means (not IQMs), IQMs have tend to have lower variance than standard means, and so may inflate effect sizes compared with other measures of central tendency.
+In response to this concern, we generated two-tailed 95% t-confidence intervals of the standard mean.
+These new CIs do not alter the main results of the paper.
 
-Figure 6 becomes:
-sigma, mean(absolute error) +/- 95%t-ci
+Fig 6 becomes:
+sigma, mean(absolute error)+/-95%t-ci
 0.05,  0.04 +/- (0.004)
@@
 
-Figure 8 becomes:
-graph type, mean(error) +/- 95%t-ci
+Fig 8 becomes:
+graph type, mean(error)+/-95%t-ci
 area,  -0.01 +/- (0.006)
@@
 
-Figure 11 becomes:
-error type, mean(absolute error) +/- 95%t-ci
+Fig 11 becomes:
+error type, mean(absolute error)+/-95%t-ci
 w/r/t OLS fit, 0.4 +/- 0.009
@@
 ...and so on.
-While estimates are noisier under this procedure, the results remain unchanged: monotonic increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts but not other chart types, participants are closer to the robust fit than the OLS fit, etc.
+While effect sizes are smaller, the results remain unchanged: increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts, estimations closer to the robust fit than the OLS fit, etc.
 
@@
 We will include or clarify study details asked for by reviewers, including the number of exclusions, and re-structuring of the methods section to make the participant information clearer.
+Though we did submit our study data as a supplement, on acceptance, we would make public our full repository containing data tables, analyses, experimental stimuli, and instruments.
 
-Additionally, though we did submit our study data as supplemental material, on acceptance, we would make public our full (currently private) repository containing not only data tables and analyses, but also experimental stimuli and instruments.
-In the main body of the paper, we will add figures with more example stimuli.
-We can also include additional figures (such as violin plots of per-subject IQMs, or other non-parametric visualizations of effect size) in the online supplement.
-
-As an alternative to our study design, both R4 and R5 mention un-constrained input of trend lines (which we had also initially considered).
+As an alternative study design, both R4 and R5 mention un-constrained input of trend lines (which we had also initially considered).
 Our design is more concerned with fitting a given model (linear or otherwise) than a freeform "sketch" of a fit, which may or may not correspond to any of the categories of provided models -- and may also suffer from additional noise such as drawing "jitter" or "satisficing" by subjects due to the higher cost of revising drawings.
 For the targeted research questions in our study, we believe our more parsimonious study design is justified.
-However, we know of others actively researching the design and implementation of experimentally tractable freeform solicitation techniques (for instance, the “draw your guess” NYT example mentioned by R5).
-We will clarify the distinction between these two types of visual regression, citing work in this area that has been published since our initial submission.
+However, we know of others actively researching the design of experimentally tractable freeform solicitation techniques (e.g., the “draw your guess” NYT example mentioned by R5).
+We will clarify the distinction between these types of task, citing work in this area that has been published since our submission.
 
-3) Better situate the work in the CHI community, and better communicate its utility:
+3) Better situate the work in the CHI community:
 
-While many graphical perception works have had a welcome reception at CHI (e.g., [10, 17]), we believe that we can improve the perceived fit of this work at CHI (as per R5) through the inclusion of an explicit implications for design subsection of the conclusion.
-We believe our results offer guidance to designers in information visualization.
-In particular, they suggest that area charts (which are a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends.
-Similarly, they suggest that standard trend lines (which are OLS lines by default in programs like Excel and other common analysis tools) may not appear to be credible if there are many outliers in the data; in such cases, it appears that viewers can be trusted to perform the robust fits visually.
-
-This issue of trust in the viewer (or the stats!) to be "well behaved" when it comes to estimating statistical properties, or making use of visualizations, is central to how we developed our research questions for this paper, and we will expand upon it in our introduction (particular our Anscombe quartet example, as per R5), and in scoping our claimed contributions.
+While many graphical perception works have had a welcome reception at CHI (e.g., [10, 17]), we believe that we can improve the perceived fit of this work at CHI (as per R5) through the inclusion of an implications for design subsection of the conclusion.
+We believe our results offer guidance to designers in visualization.
+In particular, they suggest that area charts (a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends.
+Similarly, they suggest that standard trend lines (which are OLS lines by default in programs like Excel and other common analysis tools) may not appear credible if there are many outliers in the data; in such cases, viewers could be trusted to perform the robust fits visually.
 

### 2016-11-20 Paper/Rebuttal.txt
Submitted version

@@
 sigma, mean(absolute error)+/-95%t-ci
-0.05,  0.04 +/- (0.004)
-0.10,  0.10 +/- (0.005)
-0.15,  0.15 +/- (0.006)
-0.20,  0.23 +/- (0.01)
+0.05,  0.04 +/- 0.004
+0.10,  0.10 +/- 0.005
+0.15,  0.15 +/- 0.006
+0.20,  0.23 +/- 0.01
 
@@
 graph type, mean(error)+/-95%t-ci
-area,  -0.01 +/- (0.006)
-line,  0.01 +/- (0.005)
-scatterplot, 0.005 +/- (0.003)
+area,  -0.01 +/- 0.006
+line,  0.01 +/- 0.005
+scatterplot, 0.005 +/- 0.003
 

### 2016-11-21 Paper/Rebuttal.txt
Minor edits.

@@
 
-While we reiterate that the p-values reported in the paper come from standard means (not IQMs), IQMs have tend to have lower variance than standard means, and so may inflate effect sizes compared with other measures of central tendency.
+While we reiterate that the p-values reported in the paper come from standard means (not IQMs), IQMs tend to have lower variance than standard means, and so may inflate effect sizes compared with other measures of central tendency.
 In response to this concern, we generated two-tailed 95% t-confidence intervals of the standard mean.
@@
 In particular, they suggest that area charts (a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends.
-Similarly, they suggest that standard trend lines (which are OLS lines by default in programs like Excel and other common analysis tools) may not appear credible if there are many outliers in the data; in such cases, viewers could be trusted to perform the robust fits visually.
+Similarly, they suggest that standard trend lines (which are OLS lines by default in programs like Excel and other common analysis tools) may not appear credible if there are many outliers in the data; in such cases, viewers may be better off assessing trends visually.
 

### 2017-01-02 Paper/proceedings.tex
Draft final submission.

@@
 %% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-% \toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is      granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+
+ \toappear{\scriptsize Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
 Copyrights for components of this work owned by others than ACM must be honored.
@@
 \\
-% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
-% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
+ 	{\emph{CHI 2017, May 6-11, 2017, Denver, CO, USA.} } \\
+ 	Copyright \copyright~2017 ACM ISBN 978-1-4503-4655-9/17/05\ ...\$15.00. \\
+ 	http://dx.doi.org/10.1145/3025453.3025922}
 % DOI string from ACM form confirmation}
@@
 \def\plaintitle{Regression by Eye: Estimating Trends\\in Bivariate Visualizations}
-\def\plainauthor{Removed for Review}
+\def\plainauthor{Michael Correll and Jeffrey Heer}
 \def\emptyauthor{}
@@
 \author{%
-%  \alignauthor{Michael Correll\\
-%    \affaddr{University of Washington}\\
-%    \email{mcorrell@cs.washington.edu}}\\
-%  \alignauthor{Jeffrey Heer\\
-%    \affaddr{University of Washington}\\
-%    \email{jheer@cs.washington.edu}}\\
+  \alignauthor{Michael Correll\\
+    \affaddr{University of Washington}\\
+    \email{mcorrell@cs.washington.edu}}\\
+  \alignauthor{Jeffrey Heer\\
+    \affaddr{University of Washington}\\
+    \email{jheer@cs.washington.edu}}\\
 }
@@
 We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations.
-We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models.
+We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within-the-bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models.
 % In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge.
@@
 In this section, we describe experimental design aspects shared across all experiments.
+Data tables, stimuli, and experimental instruments are available online at \url{https://github.com/uwdata/trend-bias}.
 
-\subsection{Participants}
-
-We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
-For validation, additional trials, without ambiguities such as outliers or design differences, contained the actual (OLS) trend line, which subjects then needed to simply match.
-We excluded the results of participants with average error greater than 0.2 (20 times the step size of the slider) on these validation questions.
-Across the three reported experiments and pilots, we performed 7 such exclusions.
-We recruited additional participants to replace these excluded subjects.
-Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.
-
-We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $M_{age}= 33.2$, $SD_{age}=8.8$).
-Across the subject pool, 10 reported having graduate or professional degrees, 69 college degrees, 40 at least some college, and 25 high school diplomas.
-After completing the main experimental task, participants were asked to self-assess their familiarity with charts and graphs on a 5 point Likert scale.
-The plurality (64) rated themselves as ``3.
-Some familiarity,'' and none rated themselves with the maximum rating of ``5.
-A great deal of experience.''
+In these experiments, we report effect size using the interquartile mean (or midmean) of absolute error.
+The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions.
+Cleveland \& McGill~\cite{cleveland1984graphical} used the interquartile mean to provide a more robust measure of central tendency for responses from graphical perception studies, where participant error can create long tailed distributions of error.
+This is especially the case for crowdsourced studies, where correcting for data quality issues is perceived as more difficult than in laboratory settings~\cite{heer2010crowdsourcing}.
+However, the use of interquartile means can violate the assumptions involved in performing null hypothesis significance tests (as sample means can have non-normal distributions).
+Our statistical analyses were therefore performed using standard group means.
 
@@
 We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand.
-We avoided more expressive designs (such as freeform drawing) in order to constrain the model space and to reduce the impact of factors such as noise from motor movements or heterogeneous input devices.
+While other studies investigate the use of more expressive designs (such as free-form drawing)~\cite{kim2017}, we sought to disambiguate the model selection task (e.g., the selection of a a linear or non-linear model) from the model fitting task (adjusting the parameters of the selected model).
+Our design also reduces impact of factors such as noise from motor movements or heterogeneous input devices.
 
@@
 These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.
+
+\subsection{Participants}
+
+We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%.
+For validation, additional trials, without ambiguities such as outliers or design differences, contained the actual (OLS) trend line, which subjects then needed to simply match.
+We excluded responses from participants with high average error (greater than 0.2) on these validation stimuli.
+Across the three reported experiments and pilots, we performed 7 such exclusions.
+We recruited additional participants to replace these excluded subjects.
+Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.
+
+We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $M_{age}= 33.2$, $SD_{age}=8.8$).
+Across the subject pool, 10 reported having graduate or professional degrees, 69 college degrees, 40 at least some college, and 25 high school diplomas.
+After completing the main experimental task, participants were asked to self-assess their familiarity with charts and graphs on a 5 point Likert scale.
+The plurality (64) rated themselves as ``3.
+Some familiarity,'' and none rated themselves with the maximum rating of ``5.
+A great deal of experience.''
 
@@
 
-In this experiment, the primary measure of performance was the interquartile mean (or midmean) of absolute error.
-The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions.
-Cleveland \& McGill~\cite{cleveland1984graphical} used the interquartile mean to provide a more robust measure of central tendency for responses from graphical perception studies, where participant error can create long tailed distributions of error.
-This is especially the case for crowdsourced studies, where correcting for data quality issues is perceived as more difficult than in laboratory settings~\cite{heer2010crowdsourcing}.
-
 Our results support our first hypothesis: \textbf{larger residuals result in less accuracy at regression by eye}.
@@
 
-%The average absolute error of slope estimation across all conditions had an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slopes.
 Prior work in graphical perception often measures performance as the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as $\log_2 (|\text{error}| + \frac{1}{8}$).
@@
 
-``Within the bar'' bias is a known perceptual bias involving bar charts, in which points contained in the visual area of the glyph of the bar are deemed likelier than points outside of the glyph.
+``Within-the-bar'' bias is a known perceptual bias involving bar charts, in which points contained in the visual area of the glyph of the bar are deemed likelier than points outside of the glyph.
 Newman \& Scholl~\cite{newman2012bar} encountered this bias for a sampling task: ``how likely is this point to have been drawn from the distribution represented by this bar?'' Correll \& Gleicher~\cite{correll2014error} likewise encountered this bias for inferential tasks: ``how likely is the population mean to take a particular value, given the sample represented by this bar?''
 
-We hypothesized that this bias would likewise occur in regression by eye when using visually asymmetric visualizations such as area charts.
+We hypothesized that this bias would likewise occur in regression by eye when using visually asymmetric visualizations such as area charts: a ``within-the-area'' bias.
 The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa.
 We therefore designed this experiment to elicit estimates of the y-intercepts of trends.
-A within-the-bar bias would then appear as systematic under-estimation of intercept in area charts.
+A within-the-area bias would then appear as systematic under-estimation of intercept in area charts.
 
@@
 \begin{enumerate}
-	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends in area charts, as opposed to line graphs and scatter plots.
+	\item \textbf{Area charts would be subject to within-the-area bias.} That is, participants would estimate lower values of y-intercepts of trends in area charts, as opposed to line graphs and scatter plots.
 \end{enumerate}
@@
 Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends.
-However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values will continue) can impact how viewers categorize and utilize outliers in prediction tasks~\cite{campbell2009anchoring, ji2001culture}.
+However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that runs of high or low values will continue) can impact how viewers categorize and utilize outliers in prediction tasks~\cite{campbell2009anchoring, ji2001culture}.
 In other words, regression by eye may not uniformly weight outliers, depending on their position in the plot.
@@
 However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$, $F(1,4591)=2800$, $p<0.001$).
-Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line monotonically increased almost 7-fold, from $0.07$ with 0 outliers to $0.54$ with 15 outliers.
+Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with $0$ outliers, monotonically increasing to $0.10$ with $15$ outliers), absolute error with respect to the OLS line monotonically increased almost 7-fold, from $0.07$ with $0$ outliers to $0.54$ with $15$ outliers.
 These results indicate that, while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.
@@
 Area charts are visually asymmetrical: the area below the line is filled in, and the area above it is not.
-We found that this asymmetry creates a within the bar bias: a systematic under-estimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area.
+We found that this asymmetry creates a ``within-the-area'' bias: a systematic under-estimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area.
 Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.
@@
 Similarly, the human tendency to discount or downweight outliers may result in slow adaptation to new data that does not fit into existing patterns.
+
+\subsection{Implications for Design}
+
+The estimation of trends, both for the prediction of values, and the imputation of missing values, is a common and important task in data analysis.
+We believe that our results suggest two main recommendations for designers of information visualizations that include time series or bivariate data:
+
+\begin{enumerate}
+	\item \emph{Designers do not need to annotate bivariate visualizations with trend lines in many cases}.
+Regression by eye, without explicit trend lines, is comparable in accuracy to other types of graphical perception tasks (such as comparison of value), even for a wide class of models.
+The presence of outliers is an exception to this general advice: human estimation of trends does not give much weight to outliers.
+For data where outliers are important, and expected to contribute to important trends, designers ought to visualize the outlier sensitive models directly.
+
+	\item \emph{Designers should avoid area charts for data where the perception of trends is important.} The visual asymmetry caused by filling in the area under the line results in a corresponding asymmetry in judgment of trend: an underestimation in perceived trend.
+This within-the-area bias is undesirable in many cases, but can be countered by visual encodings (such as scatter plots and line charts) that have no such visual asymmetry.
+\end{enumerate}
 
@@
 We show that viewers without statistical training can reliably estimate both linear and non-linear trends in charts such as scatter plots and line graphs.
-However, area graphs are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations.
+However, area graphs are subject to a ``within-the-area'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations.
 Regression by eye is also less sensitive to outliers than standard least-squares regression.
@@
 
-Omitted for review.
+This work supported by a Moore Foundation Data-Driven Investigator Award.
 

### 2017-01-04 Paper/proceedings.tex
JH edits
Made changes based on JH comments, need to fix some bad widows and
orphans for paragraphs on pgs 7 and 8

@@
 Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}.
-The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}, as wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts.
+The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}: narrow aspect ratios can result in overestimation in severity of trends, and wider aspect ratios in underestimation of severity.
 Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.
@@
 While other studies investigate the use of more expressive designs (such as free-form drawing)~\cite{kim2017}, we sought to disambiguate the model selection task (e.g., the selection of a a linear or non-linear model) from the model fitting task (adjusting the parameters of the selected model).
-Our design also reduces impact of factors such as noise from motor movements or heterogeneous input devices.
+Our design also reduces the impact of factors such as noise from motor movements or heterogeneous input devices.
 
@@
 Fig.
-\ref{fig:trendtypes} shows these different trend types on the same set of residuals.
+\ref{fig:trendtypes} shows these different trend types.
 
@@
 Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}.
-Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise interactions.
+Fit type was not a significant main effect ($F(2,4554)=2.60$, $p=0.074$), and post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise interactions.
 The interquartile mean of absolute error of was $0.06$ for linear and quadratic fits, and $0.07$ for trigonometric fits (see Fig.
 \ref{fig:exp1types}).
-This result suggests that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.
+This suggests that the relative unfamiliarity of non-linear trends does not result in poorer performance at regression by eye.
 
@@
 For instance, Heer \& Bostock~\cite{heer2010crowdsourcing} report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$).
-This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with the expected accuracy in comparing individual values in visualizations.
+This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with accuracy in comparing individual values in visualizations.
 
@@
 To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$).
-A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect of outlier location on interpolative value ($F(2,3441)=0.63$, $p=0.53$).
+A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant id as a random factor, found no significant effect of outlier location on interpolative value ($F(2,3441)=0.63$, $p=0.53$).
 
@@
 
-	\item \emph{Designers should avoid area charts for data where the perception of trends is important.} The visual asymmetry caused by filling in the area under the line results in a corresponding asymmetry in judgment of trend: an underestimation in perceived trend.
+	\item \emph{Designers should avoid area charts when the perception of trends is important.} The visual asymmetry caused by filling in the area under the line results in a corresponding asymmetry in judgment of trend: an underestimation in perceived trend.
 This within-the-area bias is undesirable in many cases, but can be countered by visual encodings (such as scatter plots and line charts) that have no such visual asymmetry.
@@
 Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings could similarly impact statistical judgments.
-For instance, outliers from a source that is perceived as credible could be weighted higher than outliers with more suspect provenance.
+For instance, outliers from a credible source could be weighted higher than outliers with more suspect provenance.
 Likewise, preconceptions about the volatility of data from certain domains may color how experts estimate trends.
 Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities.
-Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization~\cite{micallef2012assessing}.
+Cognitive and perceptual biases are difficult to overcome, and may require exploration of new visualization designs~\cite{micallef2012assessing}.
 
@@
 
-This work supported by a Moore Foundation Data-Driven Investigator Award.
+This work supported by a Moore Foundation Data-Driven Discovery Investigator Award.
 

### 2017-01-06 Paper/proceedings.tex
Changed keywords to semicolon delimited.
Fixed typos

@@
 \def\emptyauthor{}
-\def\plainkeywords{Information Visualization, Graphical Perception, Regression}
+\def\plainkeywords{Information Visualization; Graphical Perception; Regression}
 \def\plaingeneralterms{Visualization}
@@
 We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand.
-While other studies investigate the use of more expressive designs (such as free-form drawing)~\cite{kim2017}, we sought to disambiguate the model selection task (e.g., the selection of a a linear or non-linear model) from the model fitting task (adjusting the parameters of the selected model).
+While other studies investigate the use of more expressive designs (such as free-form drawing)~\cite{kim2017}, we sought to disambiguate the model selection task (e.g., the selection of a linear or non-linear model) from the model fitting task (adjusting the parameters of the selected model).
 Our design also reduces the impact of factors such as noise from motor movements or heterogeneous input devices.
@@
 
-This work supported by a Moore Foundation Data-Driven Discovery Investigator Award.
+This work was supported by a Moore Foundation Data-Driven Discovery Investigator Award.
 
